---
title: "Indiana Advanced Industries"
subtitle: "Comprehensive Analysis of Employment, Wages, and Economic Growth"
date: today
date-format: "MMMM D, YYYY"
author: "Riley Hudelson-Zipper"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    theme: 
      - cosmo
      - custom.scss
    fontsize: 1.1em
    fig-width: 14
    fig-height: 9
    fig-align: center
    embed-resources: true
    smooth-scroll: true
    link-external-newwindow: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false

library(tidyverse)
library(scales)
library(knitr)
library(kableExtra)
library(plotly)
library(htmlwidgets)
library(DT)
library(janitor)
library(readxl)

# Load processed data
output_dir <- "outputs_20251104"  # Main analysis output directory
output_dir_occ <- "outputs_occupation_20251104"  # Occupation analysis output directory
load(file.path(output_dir, "processed_data_jobs_gdp.RData"))

viz_dir <- file.path(output_dir, "visualizations")
viz_path <- "outputs_20251104/visualizations"

# Define color palette
initiative_colors <- c(
  "Advanced & Traded Industries" = "#1565C0",
  "AgriNovus" = "#2E7D32",
  "BioCrossroads" = "#6A1B9A",
  "Conexus - Manufacturing" = "#D84315",
  "Conexus - Logistics" = "#BF360C",
  "TechPoint" = "#F57C00",
  "Healthcare" = "#c92f6c",
  "Finance & Insurance" = "#19909a",
  "Total Employment" = "#424242"
)

# Helper function for filtering to NAICS detail level
# AgriNovus uses display_level == 3 for NAICS detail, others use display_level == 2
filter_naics_detail <- function(df) {
  df %>%
    filter(
      (initiative == "AgriNovus" & display_level == 3) |
      (initiative != "AgriNovus" & display_level == 2)
    )
}

# Helper function for beautiful tables
create_table <- function(df, title = NULL, striped = TRUE) {
  df %>%
    kable(
      format = "html",
      caption = title,
      digits = 1,
      format.args = list(big.mark = ","),
      align = "l",
      escape = FALSE
    ) %>%
    kable_styling(
      bootstrap_options = c(
        if(striped) "striped" else NULL, 
        "hover", 
        "condensed", 
        "responsive"
      ),
      full_width = FALSE,
      position = "left",
      font_size = 13
    ) %>%
    row_spec(0, bold = TRUE, background = "#f8f9fa") %>%
    as.character() %>%
    htmltools::HTML()
}
```

::: {.callout-important icon="false" appearance="simple"}
## 📊 Executive Summary

Indiana's advanced industries represent a critical component of the state's economic competitiveness. This comprehensive analysis examines **eight key initiatives** across employment trends, wage dynamics, regional concentrations, and economic productivity from **`r min(jobs_data$year)` through `r recent_year`**.

### Key Takeaways

```{r executive_summary_metrics}
#| output: asis

# Calculate headline metrics
total_advanced_jobs <- jobs_data %>%
  filter(display_level == 0, year == recent_year, 
         geo_area == "Indiana", initiative != "Total Employment") %>%
  summarise(total = sum(jobs, na.rm = TRUE)) %>%
  pull(total)

total_employment <- jobs_data %>%
  filter(display_level == 0, year == recent_year, 
         geo_area == "Indiana", initiative == "Total Employment") %>%
  pull(jobs)

advanced_share <- (total_advanced_jobs / total_employment) * 100

# Top initiative by employment
top_employer <- jobs_data %>%
  filter(display_level == 0, year == recent_year, 
         geo_area == "Indiana", initiative != "Total Employment") %>%
  arrange(desc(jobs)) %>%
  slice(1)

# Fastest growing initiative
fastest_growing <- jobs_growth %>%
  filter(display_level == 0, year == recent_year, 
         geo_area == "Indiana", initiative != "Total Employment",
         !is.na(cagr_2yr)) %>%
  arrange(desc(cagr_2yr)) %>%
  slice(1)

# Highest wage initiative
highest_wage <- wage_data %>%
  filter(display_level == 0, year == recent_year, 
         geo_area == "Indiana", initiative != "Total Employment",
         !is.na(wages)) %>%
  arrange(desc(wages)) %>%
  slice(1)

# Average wage premium
avg_wage_all <- wage_data %>%
  filter(display_level == 0, year == recent_year, 
         geo_area == "Indiana", initiative == "Total Employment") %>%
  pull(wages)

avg_wage_advanced <- wage_data %>%
  filter(display_level == 0, year == recent_year, 
         geo_area == "Indiana", initiative != "Total Employment") %>%
  left_join(
    jobs_data %>%
      filter(display_level == 0, year == recent_year, 
             geo_area == "Indiana") %>%
      select(initiative, jobs),
    by = "initiative"
  ) %>%
  summarise(avg = weighted.mean(wages, jobs, na.rm = TRUE)) %>%
  pull(avg)

wage_premium <- ((avg_wage_advanced - avg_wage_all) / avg_wage_all) * 100

cat(sprintf("
- **%s** is the largest initiative with **%s jobs**
- **%s** shows the strongest growth at **%.1f%% 2-Year CAGR**
- **%s** offers the highest wages at **%s annually**
- Advanced industries pay **%.0f%% more** than the state average
- **Significant regional variation** exists across Indiana's metro areas
", 
top_employer$initiative,
comma(round(top_employer$jobs, 0)),
fastest_growing$initiative,
as.numeric(fastest_growing$cagr_2yr),
highest_wage$initiative,
dollar(highest_wage$wages, accuracy = 1),
as.numeric(wage_premium)
))
```
:::

# Indiana State Performance {#sec-indiana}

## Employment Overview

Indiana's advanced industries form a diverse ecosystem spanning manufacturing, life sciences, agriculture innovation, energy/logistics, and technology. Understanding the employment composition across these initiatives is fundamental to assessing the state's economic strengths.

```{r indiana_employment_table}
#| tbl-cap: "Indiana Employment by Initiative"

# Get total employment for Indiana
total_employment <- jobs_data %>%
  filter(
    display_level == 0,
    year == recent_year,
    geo_area == "Indiana",
    initiative == "Total Employment"
  ) %>%
  pull(jobs)

# Prepare a clean lookup table for the growth rates
growth_rates <- jobs_growth %>%
  filter(
    display_level == 0,
    year == recent_year,
    geo_area == "Indiana",
    initiative != "Total Employment"
  ) %>%
  select(initiative, cagr_2yr)

# Now, build the main table using a join
indiana_jobs <- jobs_data %>%
  filter(
    display_level == 0, 
    year == recent_year,
    geo_area == "Indiana", 
    initiative != "Total Employment"
  ) %>%
  arrange(desc(jobs)) %>%
  # Join the growth rate data using 'initiative' as the key
  left_join(growth_rates, by = "initiative") %>%
  # Now, mutate and format the columns with clean data
  mutate(
    `Total Jobs` = scales::comma(round(jobs, 0)),
    `Share of Total Employment` = sprintf("%.1f%%", (jobs / total_employment) * 100),
    `3-Yr CAGR` = sprintf("%.1f%%", cagr_2yr)
  ) %>%
  select(
    Initiative = initiative, 
    `Total Jobs`,
    `Share of Total Employment`, 
    `3-Yr CAGR`
  )

# This function should now receive a correctly formed data frame
create_table(indiana_jobs, striped = TRUE)

```

::: column-page
```{r indiana_jobs_viz, fig.height=7}
#| fig-cap: "Total employment varies significantly across advanced industry initiatives, reflecting Indiana's diverse economic base. Manufacturing-intensive sectors dominate the employment landscape."

knitr::include_url(file.path(viz_path, "01_indiana_jobs_by_initiative.html"), height = "400px")
```
:::

::: {.callout-note icon="false"}
## 💡 Analysis

The employment distribution reveals Indiana's **manufacturing heritage** alongside emerging sectors. The dominance of traditional advanced industries reflects decades of industrial investment, while smaller but growing initiatives in life sciences and technology represent strategic diversification efforts. This portfolio approach provides economic resilience while positioning Indiana for future growth in high-value sectors.
:::

## Growth Dynamics

Growth rates reveal which initiatives are expanding and which face headwinds. The compound annual growth rate (CAGR) smooths out short-term volatility to show sustained trends.

```{r indiana_growth_table}
#| tbl-cap: "Indiana Job Growth Metrics by Initiative"

indiana_growth <- jobs_growth %>%
  filter(display_level == 0, year == recent_year, 
         geo_area == "Indiana", initiative != "Total Employment",
         !is.na(cagr_2yr)) %>%
  left_join(
    jobs_data %>%
      filter(display_level == 0, year == recent_year, 
             geo_area == "Indiana") %>%
      select(initiative, current_jobs = jobs),
    by = "initiative"
  ) %>%
  arrange(desc(cagr_2yr)) %>%
  mutate(
    `Current Jobs` = comma(round(current_jobs, 0)),
    `YoY Growth` = sprintf("%.1f%%", yoy_growth),
    `2-Year CAGR` = sprintf("%.1f%%", cagr_2yr),
    `4-Year CAGR` = if_else(!is.na(cagr_4yr), 
                            sprintf("%.1f%%", cagr_4yr), 
                            "—")
  ) %>%
  select(Initiative = initiative, `Current Jobs`, 
         `YoY Growth`, `2-Year CAGR`, `4-Year CAGR`)

create_table(indiana_growth)
```

::: column-page
```{r indiana_growth_viz, fig.height=7}
#| fig-cap: "Growth rate disparities across initiatives indicate shifting economic priorities and market dynamics. Positive growth signals expanding opportunities, while contraction suggests structural challenges or market saturation."

knitr::include_url(file.path(viz_path, "02_indiana_growth_by_initiative.html"), height = "400px")
```
:::

## Wage Analysis & Economic Value

Wage levels reflect the economic value and skill requirements of the advanced industry sectors. Higher wages typically correlate with greater worker productivity, advanced skills, and capital intensity.

```{r indiana_wage_table}
#| tbl-cap: "Indiana Average Annual Wages by Initiative"

avg_state_wage <- wage_data %>%
  filter(display_level == 0, year == recent_year, 
         geo_area == "Indiana", initiative == "Total Employment") %>%
  pull(wages)

indiana_wages <- wage_data %>%
  filter(display_level == 0, year == recent_year, 
         geo_area == "Indiana", initiative != "Total Employment",
         !is.na(wages)) %>%
  left_join(
    jobs_data %>%
      filter(display_level == 0, year == recent_year, 
             geo_area == "Indiana") %>%
      select(initiative, jobs),
    by = "initiative"
  ) %>%
  arrange(desc(wages)) %>%
  mutate(
    `Average Wage` = dollar(wages, accuracy = 1),
    `vs. State Avg` = sprintf("%+.0f%%", ((wages - avg_state_wage) / avg_state_wage) * 100),
    `Total Jobs` = comma(round(jobs, 0)),
    `Total Payroll` = dollar((wages * jobs) / 1e9, accuracy = 0.1, suffix = "B")
  ) %>%
  select(Initiative = initiative, `Average Wage`, `vs. State Avg`, 
         `Total Jobs`, `Total Payroll`)

create_table(indiana_wages)
```


```{r indiana_wage_viz, fig.height=7}
#| fig-cap: "Wage levels vary substantially across advanced industries, reflecting differences in skill requirements, capital intensity, and market dynamics. Higher-wage initiatives offer greater economic opportunity for Indiana workers."

knitr::include_url(file.path(viz_path, "04_indiana_wages_by_initiative.html"), height = "600px")

```


```{r wage_percentile_table}
#| tbl-cap: "Wage Distribution Across Initiatives (Indiana)"

wage_percentiles <- wage_data %>%
  filter(year == recent_year,
         geo_area == "Indiana", !is.na(wages)) %>%
  filter_naics_detail() %>%
  group_by(initiative) %>%
  summarise(
    `10th Percentile` = dollar(quantile(wages, 0.10, na.rm = TRUE), accuracy = 1),
    `25th Percentile` = dollar(quantile(wages, 0.25, na.rm = TRUE), accuracy = 1),
    `Median` = dollar(median(wages, na.rm = TRUE), accuracy = 1),
    `75th Percentile` = dollar(quantile(wages, 0.75, na.rm = TRUE), accuracy = 1),
    `90th Percentile` = dollar(quantile(wages, 0.90, na.rm = TRUE), accuracy = 1),
    `Industries` = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(Median))

create_table(wage_percentiles)
```

```{r in_us_wage_comparison}
#| tbl-cap: "Indiana vs. US Average Wages by Initiative"

wage_comparison <- wage_data %>%
  filter(display_level == 0, year == recent_year,
         initiative != "Total Employment", !is.na(wages),
         geo_area %in% c("Indiana", "United States")) %>%
  select(geo_area, initiative, wages) %>%
  pivot_wider(names_from = geo_area, values_from = wages) %>%
  mutate(
    `Indiana Wage` = dollar(`Indiana`, accuracy = 1),  # Add backticks here
    `US Wage` = dollar(`United States`, accuracy = 1),  # And here
    `Difference` = dollar(`Indiana` - `United States`, accuracy = 1),  # And here
    `% Difference` = sprintf("%+.1f%%", ((`Indiana` - `United States`) / `United States`) * 100),  # And here
    Gap = ifelse(`Indiana` < `United States`, "Below", "Above")  # And here
  ) %>%
  arrange(desc(`Indiana`)) %>%  # And here
  select(Initiative = initiative, `Indiana Wage`, `US Wage`, 
         `Difference`, `% Difference`, Gap)

create_table(wage_comparison)
```

## Employment Trends Over Time

Historical employment trends reveal how Indiana's advanced industries have evolved through economic cycles, technological change, and market shifts.

::: column-page
```{r indiana_timeseries_viz, fig.height=8}
#| fig-cap: "Historical employment trajectories show how each initiative has responded to economic conditions, technological change, and market dynamics. Divergent patterns reveal differential resilience and growth potential. Dashed lines indicate projected employment."

knitr::include_url(file.path(viz_path, "07_jobs_time_series_by_geography.html"), height = "950px")

```
:::

```{r indiana_timeseries_table}
#| tbl-cap: "Employment Change Over Analysis Period: Indiana vs United States"

# Define analysis period (excludes projection years)
start_year <- 2019
end_year <- recent_year  # Dynamic: uses recent_year from loaded data

# Get Indiana data - filter to actual data years only
indiana_change <- jobs_data %>%
  filter(display_level == 0, geo_area == "Indiana",
         initiative != "Total Employment",
         year >= start_year, year <= end_year) %>%
  group_by(initiative) %>%
  arrange(year) %>%
  summarise(
    Start_Year = first(year),
    End_Year = last(year),
    IN_Start_Jobs = first(jobs),
    IN_End_Jobs = last(jobs),
    .groups = "drop"
  )

# Get US data - filter to actual data years only
us_change <- jobs_data %>%
  filter(display_level == 0, geo_area == "United States",
         initiative != "Total Employment",
         year >= start_year, year <= end_year) %>%
  group_by(initiative) %>%
  arrange(year) %>%
  summarise(
    US_Start_Jobs = first(jobs),
    US_End_Jobs = last(jobs),
    .groups = "drop"
  )

# Combine and format
time_change <- indiana_change %>%
  left_join(us_change, by = "initiative") %>%
  mutate(
    Period = sprintf("%d–%d", Start_Year, End_Year),
    `IN Starting Employment` = comma(round(IN_Start_Jobs, 0)),
    `IN Current Employment` = comma(round(IN_End_Jobs, 0)),
    `IN Change` = comma(round(IN_End_Jobs - IN_Start_Jobs, 0)),
    `IN % Change` = sprintf("%.1f%%", ((IN_End_Jobs - IN_Start_Jobs) / IN_Start_Jobs) * 100),
    `US Starting Employment` = comma(round(US_Start_Jobs, 0)),
    `US Current Employment` = comma(round(US_End_Jobs, 0)),
    `US Change` = comma(round(US_End_Jobs - US_Start_Jobs, 0)),
    `US % Change` = sprintf("%.1f%%", ((US_End_Jobs - US_Start_Jobs) / US_Start_Jobs) * 100)
  ) %>%
  select(Initiative = initiative, Period, 
         `IN Starting Employment`, `IN Current Employment`, 
         `IN Change`, `IN % Change`,
         `US Starting Employment`, `US Current Employment`,
         `US Change`, `US % Change`)

create_table(time_change)
```

```{r}
#| tbl-cap: "Employment Volatility by Initiative"

# Filter to actual data years only (start_year through recent_year)
volatility <- jobs_growth %>%
  filter(display_level == 0, geo_area == "Indiana",
         initiative != "Total Employment",
         year >= start_year, year <= recent_year,  # Exclude projection years
         !is.na(yoy_growth)) %>%
  group_by(initiative) %>%
  summarise(
    mean_growth = mean(yoy_growth, na.rm = TRUE),
    std_dev = sd(yoy_growth, na.rm = TRUE),
    min_growth = min(yoy_growth, na.rm = TRUE),
    max_growth = max(yoy_growth, na.rm = TRUE),
    cov = sd(yoy_growth, na.rm = TRUE) / abs(mean(yoy_growth, na.rm = TRUE)),
    years = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(cov)) %>%
  mutate(
    `Mean YoY Growth` = sprintf("%.1f%%", mean_growth),
    `Std Dev` = sprintf("%.1f pp", std_dev),
    `Min YoY Growth` = sprintf("%.1f%%", min_growth),
    `Max YoY Growth` = sprintf("%.1f%%", max_growth),
    `Coefficient of Variation` = sprintf("%.2f", cov),
    `Years Measured` = as.character(years)
  ) %>%
  select(Initiative = initiative, `Mean YoY Growth`, `Std Dev`, 
         `Min YoY Growth`, `Max YoY Growth`, `Coefficient of Variation`, `Years Measured`)

create_table(volatility)
```

```{r}
#| output: asis

# Extract values for programmatic insertion
# Filter to actual data years only (exclude projections)
vol_data <- jobs_growth %>%
  filter(display_level == 0, geo_area == "Indiana",
         initiative != "Total Employment",
         year >= start_year, year <= recent_year,  # Exclude projection years
         !is.na(yoy_growth)) %>%
  group_by(initiative) %>%
  summarise(
    mean_growth = mean(yoy_growth, na.rm = TRUE),
    std_dev = sd(yoy_growth, na.rm = TRUE),
    min_growth = min(yoy_growth, na.rm = TRUE),
    max_growth = max(yoy_growth, na.rm = TRUE),
    cov = sd(yoy_growth, na.rm = TRUE) / abs(mean(yoy_growth, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  arrange(desc(cov))

# Get specific initiative data
conexus_mfg <- vol_data %>% filter(initiative == "Conexus - Manufacturing")
conexus_log <- vol_data %>% filter(initiative == "Conexus - Logistics")
ati <- vol_data %>% filter(initiative == "Advanced & Traded Industries")
techpoint <- vol_data %>% filter(initiative == "TechPoint")
agrinou <- vol_data %>% filter(initiative == "AgriNovus")
bio <- vol_data %>% filter(initiative == "BioCrossroads")

cat(sprintf("
- **Conexus - Manufacturing** (CoV = %.2f) shows volatility typical of manufacturing sectors, with employment swings ranging from %.1f%% to +%.1f%% and average growth of %.1f%%. This reflects sensitivity to economic cycles in advanced manufacturing.

- **Conexus - Logistics** (CoV = %.2f) exhibits volatility in logistics and distribution, with swings from %.1f%% to +%.1f%% and average growth of %.1f%%.

- **Advanced & Traded Industries** (CoV = %.2f) exhibits moderate-high volatility, typical of advanced manufacturing and other traded sectors exposed to national and global demand fluctuations.

- **TechPoint, AgriNovus, and BioCrossroads** (CoV ≈ %.2f-%.2f) demonstrate more stable growth patterns. These sectors show consistent expansion with lower cyclical sensitivity, suggesting structural growth drivers rather than cyclical fluctuations.

The standard deviation column shows the typical year-to-year swing in percentage points. For example, Conexus - Manufacturing typically varies by ±%.1f percentage points around its mean growth rate of %.1f%%, while BioCrossroads varies by ±%.1f points around %.1f%% average growth.
",
conexus_mfg$cov,
conexus_mfg$min_growth, conexus_mfg$max_growth, conexus_mfg$mean_growth,
conexus_log$cov,
conexus_log$min_growth, conexus_log$max_growth, conexus_log$mean_growth,
ati$cov,
min(bio$cov, agrinou$cov, techpoint$cov), max(bio$cov, agrinou$cov, techpoint$cov),
conexus_mfg$std_dev, conexus_mfg$mean_growth,
bio$std_dev, bio$mean_growth
))
```

# Geographic Distribution {#sec-geography}

## Metropolitan Area Concentration

Advanced industry employment is not uniformly distributed across Indiana. Metro areas vary significantly in their industrial composition, scale, and specialization patterns.

```{r metro_employment_table}
#| tbl-cap: "Employment by Initiative, Indiana and Metros, 2023"

# Get total employment for each geography
total_employment <- jobs_data %>%
  filter(display_level == 0, year == recent_year,
         initiative == "Total Employment") %>%
  select(geo_area, total_jobs = jobs)

# Get all metros plus Indiana, pivoted by initiative with percentages
all_metros_initiatives <- jobs_data %>%
  filter(display_level == 0, year == recent_year, 
         initiative != "Total Employment") %>%
  filter(geo_type %in% c("Metro", "State")) %>%
  select(geo_area, geo_type, initiative, jobs) %>%
  left_join(total_employment, by = "geo_area") %>%
  mutate(
    display_value = sprintf("%s (%.1f%%)", 
                           comma(round(jobs, 0)), 
                           (jobs / total_jobs) * 100)
  ) %>%
  select(geo_area, geo_type, initiative, display_value, total_jobs) %>%
  pivot_wider(names_from = initiative, values_from = display_value, values_fill = "0 (0.0%)") %>%
  mutate(
    sort_order = if_else(geo_type == "State", 0, 1),
    `Total Employment` = comma(round(total_jobs, 0))
  ) %>%
  arrange(sort_order, geo_area) %>%
  select(Geography = geo_area, `Total Employment`,
         `Advanced & Traded Industries`, AgriNovus, BioCrossroads,
         `Conexus - Manufacturing`, `Conexus - Logistics`, TechPoint)

all_metros_initiatives %>%
  kable(
    format = "html",
    align = c("l", "r", rep("r", ncol(all_metros_initiatives) - 2)),
    escape = FALSE
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "left",
    font_size = 13
  ) %>%
  row_spec(0, bold = TRUE, background = "#f8f9fa") %>%
  row_spec(1, bold = TRUE, background = "#e3f2fd") %>%  # Highlight Indiana row
  scroll_box(height = "500px") %>%
  as.character() %>%
  htmltools::HTML()

```

## Metro Area Growth Performance

Growth rates reveal which metro areas are **emerging** versus **mature** advanced industry centers. Understanding these dynamics helps target economic development resources.

```{r metro_growth_table}
#| tbl-cap: "Metro Area Growth Rankings (4-Year CAGR)"

metro_growth <- jobs_growth %>%
  filter(display_level == 0, year == recent_year, 
         geo_type == "Metro", initiative != "Total Employment",
         !is.na(cagr_4yr)) %>%
  group_by(geo_area) %>%
  summarise(
    total_jobs = sum(jobs, na.rm = TRUE),
    avg_cagr = weighted.mean(cagr_4yr, jobs, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(avg_cagr)) %>%
  head(15) %>%
  mutate(
    Rank = row_number(),
    `Current Employment` = comma(round(total_jobs, 0)),
    `4-Year CAGR` = sprintf("%.1f%%", avg_cagr)
  ) %>%
  select(Rank, `Metro Area` = geo_area, 
         `Current Employment`, `4-Year CAGR`)

create_table(metro_growth, striped = TRUE)
```

::: column-page
```{r metro_growth_viz, fig.height=12}
#| fig-cap: "Metro area growth rates vary substantially, with some showing robust expansion while others contract. The dropdown menu allows exploration of growth patterns across all initiatives. Starting and ending employment levels provide context for growth magnitude."

knitr::include_url(file.path(viz_path, "05_all_metros_growth_rates.html"), height = "800px")
```
:::

::: {.callout-note icon="false"}
## 💡 Analysis

Metro performance reveals a **dispersed growth pattern** rather than concentration in Indiana's largest cities. Some smaller metros achieve impressive percentage growth from modest bases, while larger metros show more moderate but substantial absolute growth. This suggests **opportunities exist beyond traditional economic centers**. Policy makers should consider whether rapid-growth smaller metros possess **replicable success factors** or unique circumstances. Growth leaders merit study as potential **models for other regions**.
:::

## Regional Specialization & Industry Mix

Metro areas develop distinct **industrial specializations** based on history, workforce skills, infrastructure, and institutional anchors. Understanding these patterns reveals competitive advantages.

::: column-page
```{r metro_mix_viz, fig.height=15}
#| fig-cap: "Initiative composition varies dramatically across metro areas, revealing distinct regional specializations. Stacked bars show absolute employment scale while proportions indicate relative specialization patterns."

knitr::include_url(file.path(viz_path, "06_metro_initiative_share.html"), height = "1200px")
```
:::

```{r metro_specialization_table}
#| tbl-cap: "Metro Area Dominant Initiatives (Highest Employment Share)"

metro_spec <- jobs_data %>%
  filter(display_level == 0, year == recent_year, 
         geo_type == "Metro", initiative != "Total Employment") %>%
  group_by(geo_area) %>%
  mutate(total_metro = sum(jobs, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(share = (jobs / total_metro) * 100) %>%
  group_by(geo_area) %>%
  slice_max(share, n = 1) %>%
  ungroup() %>%
  arrange(desc(total_metro)) %>%
  head(10) %>%
  mutate(
    `Dominant Initiative` = initiative,
    `Share of Metro Advanced Jobs` = sprintf("%.1f%%", share),
    `Initiative Employment` = comma(round(jobs, 0)),
    `Total Metro Advanced Jobs` = comma(round(total_metro, 0))
  ) %>%
  select(`Metro Area` = geo_area, `Dominant Initiative`, 
         `Share of Metro Advanced Jobs`, `Initiative Employment`, 
         `Total Metro Advanced Jobs`)

create_table(metro_spec)
```

::: {.callout-important icon="false"}
## 🎯 Strategic Implications

Regional specialization patterns suggest metro areas should pursue **differentiated strategies** rather than one-size-fits-all approaches. Areas with strong concentrations should **deepen their specialization** through cluster development, while more diversified metros might pursue **portfolio strategies**. Understanding dominant initiatives helps align **education/training programs** with local labor market needs and informs **site selection decisions** for firms seeking industry clusters.
:::

# Comparative Regional Analysis {#sec-comparative}

## Industry Specialization Patterns (Location Quotients)

**Location quotients (LQ)** measure regional industry concentration relative to the national baseline. An LQ \> 1 indicates specialization (higher concentration than US average), while LQ \< 1 suggests underrepresentation.

::: column-page
```{r lq_heatmap_viz, fig.height=15}
#| fig-cap: "Location quotients reveal which metro areas specialize in particular initiatives relative to the US average. Red shading (LQ > 1) indicates regional specialization, while blue (LQ < 1) shows underrepresentation. Metro areas can leverage specializations for competitive advantage."

knitr::include_url(file.path(viz_path, "09_location_quotient_heatmap.html"), height = "1200px")
```
:::

```{r lq_summary_table}
#| tbl-cap: "Metro Areas with Strongest Specializations (LQ > 1.5)"

lq_summary <- jobs_data %>%
  filter(display_level == 0, year == recent_year, 
         initiative != "Total Employment") %>%
  cross_join(
    jobs_data %>%
      filter(display_level == 0, year == recent_year, 
             geo_area == "United States", initiative == "Total Employment") %>%
      select(us_total = jobs)
  ) %>%
  left_join(
    jobs_data %>%
      filter(display_level == 0, year == recent_year, 
             geo_area == "United States", initiative != "Total Employment") %>%
      select(initiative, us_initiative_jobs = jobs),
    by = "initiative"
  ) %>%
  left_join(
    jobs_data %>%
      filter(display_level == 0, year == recent_year,
             initiative == "Total Employment") %>%
      select(geo_area, metro_total = jobs),
    by = "geo_area"
  ) %>%
  mutate(
    metro_pct = (jobs / metro_total) * 100,
    us_pct = (us_initiative_jobs / us_total) * 100,
    lq = metro_pct / us_pct
  ) %>%
  filter(geo_type == "Metro", lq > 1.5) %>%
  arrange(desc(lq)) %>%
  head(15) %>%
  mutate(
    `Location Quotient` = sprintf("%.2f", lq),
    `Metro %` = sprintf("%.1f%%", metro_pct),
    `US %` = sprintf("%.1f%%", us_pct)
  ) %>%
  select(`Metro Area` = geo_area, Initiative = initiative, 
         `Location Quotient`, `Metro %`, `US %`)

create_table(lq_summary, striped = TRUE)
```

::: {.callout-note icon="false"}
## 💡 Analysis

The location quotient analysis reveals **distinct regional specializations** that reflect historical development paths, institutional anchors (universities, research centers), natural advantages, and policy choices. High LQ values indicate **established competitive advantages** that can be leveraged for further growth. Metro areas with strong specializations should consider **cluster development strategies**, while those lacking specialization might pursue **strategic targeting** of specific initiatives aligned with workforce capabilities and infrastructure.
:::

### Indiana Statewide Specialization

How does Indiana's overall industry mix compare to the national average? The **statewide location quotient** shows which initiatives Indiana specializes in relative to the US.

::: column-page
```{r indiana_lq_viz, fig.height=7}
#| fig-cap: "Indiana's location quotient by initiative shows statewide specialization relative to the US average. LQ > 1 (red) indicates Indiana has a higher concentration than the national average; LQ < 1 (blue) indicates underrepresentation."

knitr::include_url(file.path(viz_path, "09b_indiana_statewide_lq.html"), height = "500px")
```
:::

```{r indiana_lq_table}
#| tbl-cap: "Indiana Industry Specialization Summary"

indiana_lq_summary <- jobs_data %>%
  filter(display_level == 0, year == recent_year,
         initiative != "Total Employment",
         geo_area == "Indiana") %>%
  cross_join(
    jobs_data %>%
      filter(display_level == 0, year == recent_year,
             geo_area == "United States", initiative == "Total Employment") %>%
      select(us_total = jobs)
  ) %>%
  left_join(
    jobs_data %>%
      filter(display_level == 0, year == recent_year,
             geo_area == "United States", initiative != "Total Employment") %>%
      select(initiative, us_initiative_jobs = jobs),
    by = "initiative"
  ) %>%
  cross_join(
    jobs_data %>%
      filter(display_level == 0, year == recent_year,
             geo_area == "Indiana", initiative == "Total Employment") %>%
      select(indiana_total = jobs)
  ) %>%
  mutate(
    indiana_pct = (jobs / indiana_total) * 100,
    us_pct = (us_initiative_jobs / us_total) * 100,
    lq = indiana_pct / us_pct
  ) %>%
  arrange(desc(lq)) %>%
  mutate(
    `Indiana %` = sprintf("%.2f%%", indiana_pct),
    `US %` = sprintf("%.2f%%", us_pct),
    `Location Quotient` = sprintf("%.2f", lq)
  ) %>%
  select(Initiative = initiative, `Indiana %`, `US %`, `Location Quotient`)

create_table(indiana_lq_summary, striped = TRUE)
```

## Initiative Composition Across Regions

How do metro areas compare in their initiative mix? This **faceted comparison** shows each initiative's employment share across all regions.

::: column-page
```{r initiative_faceted_viz, fig.height=18}
#| fig-cap: "Initiative-specific employment shares across all geographies reveal substantial regional variation. Each panel shows how one initiative's importance varies across metros, providing insights into regional competitive positioning."

knitr::include_url(file.path(viz_path, "10_initiative_mix_faceted.html"), height = "1400px")
```
:::

::: {.callout-note icon="false"}
## 💡 Analysis

The faceted view makes regional variation starkly apparent. Some initiatives show **high geographic concentration** (few metros dominate), while others are more **evenly distributed**. Concentrated initiatives may face **talent competition** and **supply chain bottlenecks** in specialized regions but benefit from **agglomeration economies**. Distributed initiatives suggest more **generalized capability** or **lower barriers to entry**. Understanding these patterns helps guide **recruitment and expansion strategies**.
:::

## Wage Premiums Across Regions

Do advanced industries pay similar wage premiums everywhere, or do regional labor markets differ? **Wage premium analysis** compares initiative wages to regional overall averages.

::: column-page
```{r wage_premium_viz, fig.height=18}
#| fig-cap: "Wage premiums (percentage above regional average) vary significantly across metro areas and initiatives. Positive values (blue) indicate advanced industries pay more than the regional average, while negative values (red) suggest below-average compensation. Regional labor market dynamics create substantial variation."

knitr::include_url(file.path(viz_path, "11_wage_premium_heatmap.html"), height = "1200px")
```
:::

```{r wage_premium_table}
#| tbl-cap: "Highest Wage Premiums by Metro Area and Initiative"

wage_premium_summary <- wage_data %>%
  filter(display_level == 0, year == recent_year, 
         initiative != "Total Employment", !is.na(wages)) %>%
  left_join(
    wage_data %>%
      filter(display_level == 0, year == recent_year,
             initiative == "Total Employment") %>%
      select(geo_area, total_wage = wages),
    by = "geo_area"
  ) %>%
  mutate(
    wage_premium = ((wages - total_wage) / total_wage) * 100
  ) %>%
  filter(geo_type %in% c("Metro", "State")) %>%
  arrange(desc(wage_premium)) %>%
  head(15) %>%
  mutate(
    `Initiative Wage` = dollar(wages, accuracy = 1),
    `Regional Average` = dollar(total_wage, accuracy = 1),
    `Premium` = sprintf("%+.0f%%", wage_premium)
  ) %>%
  select(`Geography` = geo_area, Initiative = initiative, 
         `Initiative Wage`, `Regional Average`, `Premium`)

create_table(wage_premium_summary)
```

## Employment Growth Trajectories

How do different regions' growth paths compare over time? **Indexed employment trends** make it easy to see relative performance since a common baseline.

::: column-page
```{r growth_trajectories_viz, fig.height=14}
#| fig-cap: "Indexed employment trends (base year = 100) reveal divergent growth trajectories across metro areas. Gray lines show individual metros while Indiana (green) and US (red) provide benchmark comparisons. Dashed lines indicate projected values. Use the dropdown to examine different initiatives."

knitr::include_url(file.path(viz_path, "12_growth_comparison_interactive.html"), height = "1200px")
```
:::

::: {.callout-note icon="false"}
## 💡 Analysis

The indexed view reveals **heterogeneous recovery and growth patterns**. Some metros show **consistent outperformance** of state and national benchmarks, suggesting strong local fundamentals or effective strategies. Others lag persistently, indicating **structural challenges** or competitive disadvantages. The **pandemic shock** created divergent paths—some metros proved resilient while others suffered lasting setbacks. These patterns should inform **targeted interventions** and help identify **best practices** from high-performing regions.
:::

# Strategic Positioning Analysis {#sec-strategic}

## Growth vs. Scale Quadrant Analysis

The **Growth-Level Quadrant** positions metro areas based on their employment growth rate versus employment share of total metro employment. This reveals strategic positioning.

::: column-page
```{r quadrant_viz, fig.height=18}
#| fig-cap: "Quadrant analysis positions metros by growth rate (vertical) and employment share (horizontal), with median values as reference lines. **Stars** (high growth, high share) are thriving clusters. **Emerging** (high growth, low share) are up-and-comers. **Mature** (low growth, high share) are established but slowing. **Declining** sectors merit attention. Use dropdown to examine different initiatives."

knitr::include_url(file.path(viz_path, "13_growth_vs_level_quadrant_interactive.html"), height = "1000px")
```
:::

```{r quadrant_table}
#| tbl-cap: "Metro Strategic Positioning Summary"

quadrant_summary <- jobs_growth %>%
  filter(display_level == 0, year == recent_year, 
         geo_type == "Metro", initiative != "Total Employment",
         !is.na(cagr_4yr)) %>%
  left_join(
    jobs_data %>%
      filter(display_level == 0, year == recent_year,
             initiative == "Total Employment") %>%
      select(geo_area, total_employment = jobs),
    by = "geo_area"
  ) %>%
  mutate(
    employment_share = (jobs / total_employment) * 100
  ) %>%
  group_by(initiative) %>%
  mutate(
    median_growth = median(cagr_4yr, na.rm = TRUE),
    median_share = median(employment_share, na.rm = TRUE),
    quadrant = case_when(
      cagr_4yr >= median_growth & employment_share >= median_share ~ "⭐ Stars",
      cagr_4yr >= median_growth & employment_share < median_share ~ "🚀 Emerging",
      cagr_4yr < median_growth & employment_share >= median_share ~ "📊 Mature",
      TRUE ~ "⚠️ Declining"
    )
  ) %>%
  ungroup() %>%
  group_by(initiative, quadrant) %>%
  summarise(n_metros = n(), .groups = "drop") %>%
  pivot_wider(names_from = quadrant, values_from = n_metros, values_fill = 0)

create_table(quadrant_summary, striped = TRUE)
```

::: {.callout-important icon="false"}
## 🎯 Strategic Implications

**Quadrant positioning** suggests differentiated strategies:

-   **Stars** warrant **sustained investment** and **cluster deepening** to maintain momentum
-   **Emerging** areas need **infrastructure and talent pipeline development** to support growth
-   **Mature** sectors require **innovation and productivity initiatives** to reignite growth
-   **Declining** areas need **transition planning** and **diversification strategies**

The quadrant analysis helps prioritize economic development resources and identify metros that could serve as models for others in similar positions.
:::

## Growth Acceleration & Deceleration

Is employment growth **speeding up or slowing down**? Comparing early-period to late-period growth rates reveals acceleration or deceleration trends.

::: column-page
```{r acceleration_viz, fig.height=15}
#| fig-cap: "Acceleration analysis compares early-period CAGR (horizontal) to late-period CAGR (vertical). Points above the diagonal line show **accelerating** growth, while points below show **decelerating** growth. This reveals momentum shifts and helps identify emerging trends. Use dropdown to examine different initiatives."

knitr::include_url(file.path(viz_path, "14_acceleration_deceleration_interactive.html"), height = "1000px")
```
:::

```{r acceleration_table}
#| tbl-cap: "Metro Areas with Greatest Growth Acceleration"

max_est_year <- 2024

accel_summary <- jobs_data %>%
  filter(display_level == 0,
         initiative != "Total Employment",
         geo_type %in% c("Metro", "State", "US")) %>%
  arrange(initiative, geo_area, year) %>%
  group_by(initiative, geo_area, geo_type) %>%
  summarise(
    years_available = n(),
    min_year = min(year),
    max_year = max_est_year,
    mid_year = min_year + ceiling((max_year - min_year) / 2),
    early_cagr = if_else(
      years_available >= 6,
      ((jobs[year == mid_year] / jobs[year == min_year])^(1/(mid_year - min_year)) - 1) * 100,
      NA_real_
    ),
    late_cagr = if_else(
      years_available >= 6,
      ((jobs[year == max_year] / jobs[year == mid_year])^(1/(max_year - mid_year)) - 1) * 100,
      NA_real_
    ),
    total_jobs = jobs[year == max_year],
    .groups = "drop"
  ) %>%
  filter(!is.na(early_cagr), !is.na(late_cagr),
         !is.infinite(early_cagr), !is.infinite(late_cagr)) %>%
  mutate(
    acceleration = late_cagr - early_cagr,
    trend = if_else(late_cagr > early_cagr, "Accelerating", "Decelerating")
  ) %>%
  filter(trend == "Accelerating", geo_type == "Metro") %>%
  arrange(desc(acceleration)) %>%
  head(15) %>%
  mutate(
    `Early CAGR` = sprintf("%.1f%%", early_cagr),
    `Late CAGR` = sprintf("%.1f%%", late_cagr),
    `Acceleration` = sprintf("%+.1f pp", acceleration),
    `Current Jobs` = comma(round(total_jobs, 0))
  ) %>%
  select(`Metro Area` = geo_area, Initiative = initiative,
         `Early CAGR`, `Late CAGR`, `Acceleration`, `Current Jobs`)

create_table(accel_summary)
```

::: {.callout-note icon="false"}
## 💡 Analysis

Growth acceleration patterns reveal **shifting momentum** across metro areas and initiatives. Metros showing **acceleration** may have recently implemented successful strategies, attracted key investments, or benefited from favorable market trends. Those showing **deceleration** may face emerging headwinds, saturation, or competitive pressure. Understanding these dynamics helps distinguish **durable trends** from temporary fluctuations and can inform forecasting and strategic planning.
:::

# Economic Productivity & Impact {#sec-productivity}

## GDP Growth vs. Employment Growth

The relationship between **GDP growth** and **jobs growth** reveals productivity dynamics. When GDP grows faster than employment, **productivity is improving**.

::: column-page
```{r gdp_jobs_scatter_viz, fig.height=14}
#| fig-cap: "GDP growth vs. jobs growth reveals productivity patterns across metro areas. Points above the diagonal have GDP growing faster than employment (productivity gains). Bubble size represents total jobs. The diagonal line represents proportional growth—points above show productivity improvement."

knitr::include_url(file.path(viz_path, "08_gdp_vs_jobs_growth_scatter.html"), height = "1000px")
```
:::

```{r productivity_table}
#| tbl-cap: "Geographies by Productivity and Employment Growth"

if(exists("gdp_growth") && nrow(gdp_growth) > 0) {
  productivity_summary <- gdp_growth %>%
    filter(display_level == 0, 
           geo_type %in% c("Metro", "State", "US"),
           initiative == "Total Employment") %>%
    arrange(geo_area, year) %>%  # Ensure proper ordering for lag
    group_by(geo_area) %>%  # Group so lag works within each geography
    mutate(
      jobs_growth = yoy_growth,  # Already calculated
      gdp_growth = (grp / lag(grp) - 1) * 100  # Calculate before filtering year
    ) %>%
    ungroup() %>%  # Ungroup before filtering
    filter(year == recent_year,  # NOW filter to recent year
           !is.na(yoy_growth),
           !is.na(gdp_growth),
           !is.infinite(gdp_growth)) %>%
    mutate(
      gdp_per_job = grp / jobs,
      sort_order = case_when(
        geo_area == "United States" ~ 1,
        geo_area == "Indiana" ~ 2,
        TRUE ~ 3
      )
    ) %>%
    arrange(sort_order, geo_area) %>%
    mutate(
      `Jobs Growth` = sprintf("%.1f%%", jobs_growth),
      `GDP Growth` = sprintf("%.1f%%", gdp_growth),
      `GDP per Job` = dollar(gdp_per_job, accuracy = 1),
      Year = as.character(recent_year)
    ) %>%
    select(Geography = geo_area, Year, `Jobs Growth`, `GDP Growth`,
           `GDP per Job (Labor Productivity)` = `GDP per Job`)
  
  create_table(productivity_summary, striped = TRUE)
}

```

::: {.callout-note icon="false"}
## 💡 Analysis

The productivity lens reveals **economic efficiency** beyond raw employment numbers. Regions with **strong productivity gains** (GDP growth exceeding jobs growth) may be adopting automation, improving processes, or shifting to higher-value activities. This is generally positive for **economic output and wages** but may present **workforce transition challenges**. Conversely, regions where jobs grow faster than GDP may indicate **declining productivity** or **wage pressure**, warranting investigation into competitiveness factors.
:::

# Key Findings {#sec-findings}

## Summary of Critical Insights

::: {.callout-tip icon="false"}
## 📌 Employment Landscape

```{r findings_employment}
#| output: asis

# Top 3 by employment
top3_employment <- jobs_data %>%
  filter(display_level == 0, year == recent_year, 
         geo_area == "Indiana", initiative != "Total Employment") %>%
  arrange(desc(jobs)) %>%
  head(3)

# Top 3 by growth
top3_growth <- jobs_growth %>%
  filter(display_level == 0, year == recent_year, 
         geo_area == "Indiana", initiative != "Total Employment",
         !is.na(cagr_2yr)) %>%
  arrange(desc(cagr_2yr)) %>%
  head(3)

cat("**Scale Leaders:**\n")
for(i in 1:3) {
  cat(sprintf("- **%s**: %s jobs\n\n", 
              top3_employment$initiative[i],
              comma(round(top3_employment$jobs[i], 0))))
}

cat("\n**Growth Leaders:**\n\n")
for(i in 1:3) {
  cat(sprintf("- **%s**: %.1f%% 2-Year CAGR\n\n", 
              top3_growth$initiative[i],
              top3_growth$cagr_2yr[i]))
}
```
:::

::: {.callout-tip icon="false"}
## 💰 Wage & Economic Value

```{r findings_wages}
#| output: asis

# Top 3 by wages
top3_wages <- wage_data %>%
  filter(display_level == 0, year == recent_year, 
         geo_area == "Indiana", initiative != "Total Employment",
         !is.na(wages)) %>%
  arrange(desc(wages)) %>%
  head(3)

# Calculate total payroll impact
total_payroll <- wage_data %>%
  filter(display_level == 0, year == recent_year, 
         geo_area == "Indiana", initiative != "Total Employment") %>%
  left_join(
    jobs_data %>%
      filter(display_level == 0, year == recent_year, 
             geo_area == "Indiana") %>%
      select(initiative, jobs),
    by = "initiative"
  ) %>%
  summarise(total = sum(wages * jobs, na.rm = TRUE) / 1e9) %>%
  pull(total)

cat("**Highest-Paying Initiatives:**\n\n")
for(i in 1:3) {
  cat(sprintf("- **%s**: %s average annual wage\n\n", 
              top3_wages$initiative[i],
              dollar(top3_wages$wages[i], accuracy = 1)))
}

cat(sprintf("\n**Total Annual Payroll**: %s across all advanced industries\n\n",
            dollar(total_payroll, accuracy = 0.1, suffix = "B")))
```
:::

::: {.callout-tip icon="false"}
## 🌍 Geographic Patterns

```{r findings_geography}
#| output: asis

# Top 3 metros by Advanced & Traded Industries share of total employment
top3_ati_metros <- jobs_data %>%
  filter(display_level == 0, year == recent_year, 
         geo_type == "Metro") %>%
  select(geo_area, initiative, jobs) %>%
  pivot_wider(names_from = initiative, values_from = jobs) %>%
  mutate(
    ati_share = (`Advanced & Traded Industries` / `Total Employment`) * 100
  ) %>%
  arrange(desc(ati_share)) %>%
  head(5)

# Most specialized metros
top_specialists <- jobs_data %>%
  filter(display_level == 0, year == recent_year, 
         initiative != "Total Employment") %>%
  cross_join(
    jobs_data %>%
      filter(display_level == 0, year == recent_year, 
             geo_area == "United States", initiative == "Total Employment") %>%
      select(us_total = jobs)
  ) %>%
  left_join(
    jobs_data %>%
      filter(display_level == 0, year == recent_year, 
             geo_area == "United States", initiative != "Total Employment") %>%
      select(initiative, us_initiative_jobs = jobs),
    by = "initiative"
  ) %>%
  left_join(
    jobs_data %>%
      filter(display_level == 0, year == recent_year,
             initiative == "Total Employment") %>%
      select(geo_area, metro_total = jobs),
    by = "geo_area"
  ) %>%
  mutate(
    metro_pct = (jobs / metro_total) * 100,
    us_pct = (us_initiative_jobs / us_total) * 100,
    lq = metro_pct / us_pct
  ) %>%
  filter(geo_type == "Metro") %>%
  arrange(desc(lq)) %>%
  head(5)

cat("**Highest Advanced & Traded Industries Concentration:**\n\n")
for(i in 1:5) {
  cat(sprintf("- **%s**: %.1f%% of total employment\n\n", 
              top3_ati_metros$geo_area[i],
              top3_ati_metros$ati_share[i]))
}

cat("**Strongest Specializations (Highest LQ):**\n\n")
for(i in 1:5) {
  cat(sprintf("- **%s** - %s (LQ: %.2f)\n\n", 
              top_specialists$geo_area[i],
              top_specialists$initiative[i],
              top_specialists$lq[i]))
}
```
:::

# Industry Deep Dive {#sec-industry-deepdive}

Moving beyond initiative-level aggregates, this section examines the **individual industries** that compose each advanced industry initiative. Understanding industry-level composition reveals the **specific sectors driving growth**, identifies **concentration risks**, and highlights opportunities for **targeted interventions**.

## Industry Composition Within Initiatives

Each initiative comprises multiple industries with varying employment levels. Some initiatives are **highly concentrated** in a few dominant sectors, while others exhibit more **diversified portfolios**.

::: column-page
```{r top_industries_viz, fig.height=14}
#| fig-cap: "The top 10 industries by employment within each initiative reveal concentration patterns and dominant sectors. Use the dropdown to examine different initiatives and geographies."

knitr::include_url(file.path(viz_path, "15_top_industries_by_initiative.html"), height = "1200px")
```
:::

```{r industry_concentration_summary}
#| tbl-cap: "Initiative Concentration Metrics"

# Load deep dive data if not already loaded
if(!exists("initiative_concentration")) {
  load(file.path(output_dir, "processed_data_initiative_deepdive.RData"))
}

concentration_summary <- initiative_concentration %>%
  filter(geo_area == "Indiana", year == recent_year) %>%
  arrange(desc(top5_concentration)) %>%
  mutate(
    `Top 5 Concentration` = sprintf("%.1f%%", top5_concentration),
    `Top 10 Concentration` = sprintf("%.1f%%", top10_concentration),
    `Total Industries` = comma(total_industries),
  ) %>%
  select(Initiative = initiative, `Total Industries`, 
         `Top 5 Concentration`, `Top 10 Concentration`)

create_table(concentration_summary, striped = TRUE)
```

::: {.callout-note icon="false"}
## 📊 Interpretation

**Concentration metrics** reveal strategic implications:

- **Top 5 Concentration** shows the share of initiative employment in the five largest industries. Higher values indicate concentration in fewer sectors.
- Highly concentrated initiatives face **greater sector-specific risks** but may benefit from **stronger cluster effects** and **specialized infrastructure**.
- Diversified initiatives offer **portfolio resilience** but may struggle to achieve the **critical mass** needed for cluster competitiveness.
:::

## Industry Concentration Curves

**Lorenz curves** visualize how employment is distributed across industries within each initiative. A curve closer to the diagonal indicates more equal distribution, while curves bowing toward the axes reveal concentration in just a few industries.

::: column-page
```{r concentration_curves_viz, fig.height=12}
#| fig-cap: "Lorenz curves show cumulative employment share by industry rank. Curves further from the diagonal (closer to the axes) indicate higher concentration in fewer industries. The Gini coefficient quantifies inequality (0 = perfect equality, 1 = perfect inequality)."

knitr::include_url(file.path(viz_path, "16_industry_concentration_lorenz.html"), height = "1000px")
```
:::

::: {.callout-important icon="false"}
## 🎯 Strategic Implications

The degree of industry concentration within initiatives suggests different **economic development strategies**:

- **Highly concentrated initiatives** should prioritize **cluster development** and **supply chain deepening** around dominant industries, while monitoring **concentration risk**.
- **Moderately concentrated initiatives** can pursue **adjacent diversification** by supporting industries that complement existing strengths.
- **Highly diversified initiatives** may benefit from **targeted investment** to build **critical mass** in high-potential sectors rather than spreading resources too thin.

Understanding concentration patterns helps **prioritize resources**, **design interventions**, and **manage portfolio risk** at both state and regional levels.
:::

## Growth Contributors: Which Industries Drive Change?

Not all industries contribute equally to initiative employment growth. Identifying the **specific industries** responsible for gains or losses enables **targeted policy interventions** and **resource allocation**.

::: column-page
```{r growth_contributors_viz, fig.height=14}
#| fig-cap: "Waterfall charts decompose employment change into industry-level contributions. Positive bars show industries adding jobs; negative bars indicate job losses. Use dropdowns to select geography and time period."

knitr::include_url(file.path(viz_path, "17_growth_contributors_waterfall.html"), height = "1200px")
```
:::

::: {.callout-important icon="false"}
## ❓ How to Read This Chart

This waterfall chart displays the top 10 industries contributing most to job growth within the selected initiative and geography. Industries are ranked from smallest to largest net change in jobs. 

- **Green bars** represent industries that gained jobs (positive contribution)
- **Red bars** represent industries that lost jobs (negative contribution)
- The **horizontal position** of each bar shows the cumulative job change - as you read from **bottom to top**, each bar adds to (or subtracts from) the running total
- The number on each bar shows that industry's individual job change

**Key insights**:

- Industries at the **bottom** of each chart had the largest impact on total employment growth
- The cumulative effect shows how multiple industries combine to drive overall growth
- A mix of <span style="color: green;">green</span> and <span style="color: red;">red</span> bars indicates **offsetting trends** - some industries <span style="color: green;">growing</span> while others <span style="color: red;">contract</span>

**Example**: If the bottom bar is <span style="color: green;">green</span> and shows +500 jobs and ends at 500, and the next bar is <span style="color: green;">green</span> and shows +300 jobs and ends at 800, the second industry added 300 jobs on top of the first industry's 500. If the next bar is <span style="color: red;">red</span> and shows -100, the third industry shed 100 jobs from the first two industries, for cumulative growth of 700 jobs. 
:::

```{r top_growth_contributors}
#| tbl-cap: "Top 3 Growth Contributors by Initiative (Indiana)"

if(!exists("industry_growth_contrib")) {
  load(file.path(output_dir, "processed_data_initiative_deepdive.RData"))
}

top_contributors <- industry_growth_contrib %>%
  filter(geo_area == "Indiana") %>%
  group_by(initiative) %>%
  arrange(desc(yoy_change)) %>%
  slice_head(n = 3) %>%
  ungroup() %>%
  mutate(
    `Job Change` = comma(round(yoy_change, 0)),
    `% of Initiative Growth` = sprintf("%.1f%%", contribution_pct)
  ) %>%
  select(Initiative = initiative, Industry = naics_title, 
         `Job Change`, `% of Initiative Growth`) %>%
  arrange(Initiative, desc(as.numeric(gsub(",", "", `Job Change`))))

create_table(top_contributors)
```

::: {.callout-tip icon="false"}
## 💡 Analysis

**Growth decomposition** reveals:

- A small number of industries often account for the **majority of employment change** within initiatives, suggesting that **targeted interventions** in high-impact sectors can yield disproportionate returns.
- In most geography-initiative combinations, there's a mix of red and green bars, which indicates there are **fewer than 10 industries** that grew year-over-year in that geography-initiative. 
- Industries driving growth vary by **geography and initiative**, indicating the need for **context-specific strategies** rather than one-size-fits-all approaches.
- Identifying **consistent growth contributors** across multiple periods helps distinguish **structural trends** from **cyclical fluctuations**.
- Understanding which industries drive **losses** is equally important for **transition planning** and **workforce retraining** efforts.
:::

## Wage Distribution Across Industries

While initiative-level average wages provide useful benchmarks, **substantial variation exists across component industries**. Understanding this distribution reveals opportunities to **promote high-wage sectors** and identifies industries where **wage improvements** could have significant impact.

::: column-page
```{r wage_distribution_viz, fig.height=14}
#| fig-cap: "Box plots show wage distribution within each initiative. Wider boxes indicate greater wage variation. Points beyond whiskers represent outlier industries with exceptionally high or low wages."

knitr::include_url(file.path(viz_path, "18_wage_distribution_boxplots.html"), height = "1000px")
```
:::

```{r wage_extremes_table}
#| tbl-cap: "Highest and Lowest Wage Industries by Initiative (Indiana)"

if(!exists("wage_extremes_by_init")) {
  load(file.path(output_dir, "processed_data_initiative_deepdive.RData"))
}

wage_extremes_display <- wage_extremes_by_init %>%
  filter(geo_area == "Indiana", year == recent_year) %>%
  select(initiative, category, naics_title, wages, jobs) %>%
  mutate(
    `Average Wage` = dollar(wages, accuracy = 1),
    `Employment` = comma(jobs)
  ) %>%
  select(Initiative = initiative, `Wage Category` = category, 
         Industry = naics_title, `Average Wage`, Employment) %>%
  arrange(Initiative, `Wage Category`)

create_table(wage_extremes_display)
```

::: {.callout-important icon="false"}
## 🎯 Strategic Implications

**Wage heterogeneity** within initiatives suggests several strategic opportunities:

- Initiatives with **large wage spreads** offer opportunities to **shift employment mix** toward higher-paying industries through targeted **recruitment** and **business development**.
- Understanding which specific industries pay **premium wages** helps focus **workforce development** and **talent attraction** efforts on the most impactful sectors.
- Industries with **high employment but below-average wages** represent opportunities for increasing productivity and  **quality job creation**.
- The presence of **high-wage outliers** within an initiative indicates potential for **value chain upgrading** and **innovation-driven wage growth**.
:::

## Industry Size vs. Growth: Strategic Positioning

The relationship between **current employment levels** and **growth rates** reveals strategic positioning of individual industries within each initiative.

::: column-page
```{r industry_scatter_viz, fig.height=16}
#| fig-cap: "Scatter plots position industries by current employment (bubble size) and growth rate (vertical axis). Large, fast-growing industries (top-right) are key drivers. Small, fast-growing industries (top-left) are emerging opportunities. Use dropdown to examine different initiatives."

knitr::include_url(file.path(viz_path, "19_size_vs_growth_scatter.html"), height = "1200px")
```
:::

::: {.callout-note icon="false"}
## 📈 Quadrant Interpretation

Industries fall into distinct strategic categories:

- **Stars** (Large + Fast-Growing): Dominant industries driving overall initiative growth. These merit continued support and should be protected through cluster development initiatives.
- **Emerging Opportunities** (Small + Fast-Growing): High-growth industries with potential to become significant employment drivers. These warrant investment in infrastructure and talent pipelines.
- **Mature Anchors** (Large + Slow/Negative Growth): Established industries facing headwinds. These may need productivity enhancements, innovation initiatives, or transition support.
- **Declining Sectors** (Small + Negative Growth): Industries losing jobs without significant current employment. May warrant **sunset strategies** or **repurposing resources** toward higher-growth sectors.
:::

## Regional Industry Specialization

Indiana metros don't simply replicate the state's industry mix — they develop **distinct specializations** within each initiative. **Location quotients (LQ)** identify which industries are **over-represented** in specific regions relative to the U.S. average.

::: column-page
```{r specialized_industries_viz, fig.height=16}
#| fig-cap: "Heatmaps show the top 3 most specialized industries within each initiative by metro area. Darker colors indicate stronger specialization (higher LQ). Use dropdown to examine different geographies."

knitr::include_url(file.path(viz_path, "20_specialized_industries_heatmap.html"), height = "1200px")
```
:::

```{r specialization_summary}
#| tbl-cap: "Most Specialized Industries by Geography (LQ)"

# Calculate traditional LQ for all geographies
top_specializations_all <- jobs_data %>%
  filter(year == recent_year,
         initiative != "Total Employment",
         naics_code != "000000",
         geo_area != "United States",
         !is.na(jobs)) %>%
  filter_naics_detail() %>%
  distinct(statefips, countyfips, metrofips, geo_area, initiative, 
           naics_code, naics_title, year, .keep_all = TRUE) %>%
  # Get total employment for each geography
  left_join(
    jobs_data %>%
      filter(display_level == 0, year == recent_year,
             initiative == "Total Employment") %>%
      select(statefips, countyfips, metrofips, geo_area, total_employment = jobs),
    by = c("statefips", "countyfips", "metrofips", "geo_area")
  ) %>%
  # Get US industry employment
  left_join(
    jobs_data %>%
      filter(year == recent_year,
             geo_area == "United States",
             naics_code != "000000") %>%
      filter_naics_detail() %>%
      distinct(initiative, naics_code, naics_title, .keep_all = TRUE) %>%
      select(initiative, naics_code, naics_title, us_industry_jobs = jobs),
    by = c("initiative", "naics_code", "naics_title")
  ) %>%
  # Get US total employment
  left_join(
    jobs_data %>%
      filter(display_level == 0, year == recent_year,
             geo_area == "United States", initiative == "Total Employment") %>%
      select(us_total = jobs),
    by = character()
  ) %>%
  mutate(
    local_share = (jobs / total_employment) * 100,
    us_share = (us_industry_jobs / us_total) * 100,
    location_quotient = local_share / us_share
  ) %>%
  filter(!is.na(location_quotient), 
         !is.infinite(location_quotient),
         jobs >= 100) %>%
  # Get top 5 per initiative per geography
  group_by(geo_area, initiative) %>%
  arrange(desc(location_quotient)) %>%
  slice_head(n = 5) %>%
  ungroup() %>%
  mutate(
    `Location Quotient` = round(location_quotient, 2),
    Jobs = jobs,
    `% of Total Employment` = round(local_share, 3),
    # Create sort order: Indiana first, then MSAs alphabetically
    sort_order = if_else(geo_area == "Indiana", "0", paste0("1_", geo_area))
  ) %>%
  arrange(sort_order, initiative, desc(`Location Quotient`)) %>%
  select(Geography = geo_area, Initiative = initiative, Industry = naics_title,
         `Location Quotient`, Jobs, `% of Total Employment`)

# Create interactive table with search/filter
datatable(
  top_specializations_all,
  filter = 'top',
  options = list(
    pageLength = 15,
    autoWidth = TRUE,
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel'),
    ordering = TRUE,
    order = list()  # Start with the default order from the data
  ),
  rownames = FALSE
) %>%
  formatRound(columns = c('Location Quotient', '% of Total Employment'), digits = 2) %>%
  formatCurrency(columns = 'Jobs', currency = '', digits = 0)
```

::: {.callout-important icon="false"}
## 🎯 Regional Strategy Implications

**Industry-level specialization patterns** inform differentiated regional strategies:

- Metros with **high LQ industries** possess **competitive advantages** that should be leveraged through cluster development, specialized workforce programs, and targeted marketing to related firms.
- **Comparing specializations across metros** reveals opportunities for **inter-regional collaboration** (complementary strengths) or **competitive differentiation** (similar specializations).
- Industries with **high LQ but small absolute employment** represent **niche specializations** that may offer unique competitive positioning but face scalability constraints.
- Metros lacking specialization in fast-growing industries may benefit from **strategic recruitment** efforts or **entrepreneurship support** to build new clusters.

Understanding these patterns at the industry level provides the **granular insights** needed to move from **generic strategies** to **targeted, evidence-based interventions** tailored to each region's unique industrial composition.
:::

# Business Formation & Establishment Dynamics {#sec-establishments}

Indiana's entrepreneurial ecosystem and business formation patterns provide critical insights into economic vitality and future growth potential. This section examines new establishment formation across advanced industry initiatives, distinguishing between true startups and larger qualified establishments.

Note that the two new initiatives, Finance & Insurance and Healthcare, have not yet been integrated into the business formation data.

## New Establishment Formation Overview

New establishments represent fresh economic activity and entrepreneurial dynamism. Understanding where and how new businesses form helps identify emerging clusters and entrepreneurial hotspots.

::: column-page
```{r new_establishments_timeline, fig.height=14}
#| fig-cap: "New establishment formation trends reveal entrepreneurial activity across initiatives and geographies. Use the dropdown to explore different regions. Distinct patterns emerge between true startups (1-20 employees) and larger qualified establishments (21+ employees)."

knitr::include_url("estab_output_2022/plots/A_new_establishments_timeline_interactive.html", height = "1000px")
```
:::

::: {.callout-note icon="false"}
## 💡 Analysis

Sustained formation of new establishments indicates a **healthy innovation ecosystem**, while declining trends may signal **barriers to entry** or **market saturation**. The distinction between startup-scale establishments (1-20 employees) and larger qualified establishments (21+ employees) reveals whether growth is driven by **small-scale entrepreneurship** or **larger-scale business development**.
:::

## Year-over-Year Growth in New Establishments

Examining the growth rate of new establishment formation reveals acceleration or deceleration in entrepreneurial activity.

::: column-page
```{r yoy_growth_heatmap, fig.height=14}
#| fig-cap: "Year-over-year growth rates in new establishment formation show momentum shifts across initiatives and geographies. Darker colors indicate stronger growth, revealing which sectors and regions are experiencing entrepreneurial surges."

knitr::include_url("estab_output_2022/plots/B_yoy_growth_heatmap_interactive.html", height = "1000px")
```
:::

::: {.callout-important icon="false"}
## 🎯 Strategic Implications

**Acceleration in new establishment formation** signals:

- Improving **business climate** and **entrepreneurial support infrastructure**
- Emerging **market opportunities** attracting new entrants
- Potential **cluster formation** in high-growth sectors

**Deceleration or negative growth** may indicate:

- **Maturing markets** with higher barriers to entry
- **Economic headwinds** or **policy constraints**
- **Consolidation trends** reducing net new establishment formation

Understanding these dynamics helps **target entrepreneurship support programs**, **assess economic development effectiveness**, and **identify emerging clusters** before they appear in traditional employment metrics.
:::

## Cohort Performance Analysis

Tracking establishment cohorts over time reveals survival rates and growth patterns. This longitudinal perspective distinguishes between **sustainable entrepreneurship** and **high-churn activity**.

<!-- ```{r cohort_performance_table}
#| tbl-cap: "Establishment Cohort Performance by Initiative"

cohort_data <- read_csv("estab_output_2022/tables/Q_cohort_performance_all_initiatives.csv", 
                        show_col_types = FALSE)

# Format the data using actual column names
cohort_display <- cohort_data %>%
  mutate(
    # Format year columns (no commas)
    across(matches("year", ignore.case = TRUE), ~as.character(.x)),
    # Format count/employment columns (with commas)
    across(where(is.numeric) & !matches("year|rate|growth|pct", ignore.case = TRUE), ~comma(round(.x, 0))),
    # Format rate/percentage columns (just add %)
    across(matches("rate|growth|pct", ignore.case = TRUE) & where(is.numeric), ~sprintf("%.1f%%", .x))
  ) %>%
  # Clean up column names for display
  rename_with(~str_replace_all(.x, "_", " ") %>% 
               str_to_title() %>%
               str_replace("Yr", "Year") %>%
               str_replace("Avg", "Average"))

create_table(cohort_display, striped = TRUE)
``` -->

<!-- ::: {.callout-note icon="false"}
## 📊 Cohort Insights

**Cohort analysis** reveals the long-term viability of new establishment formation:

- **Higher survival rates** indicate supportive ecosystems with adequate infrastructure, talent, and market access
- **Employment growth** in surviving establishments shows whether businesses are scaling or remaining small
- **Variation across cohorts** may reflect economic cycle impacts, with recession cohorts often showing different patterns
- **Initiative-level differences** highlight which sectors foster more durable entrepreneurship

Comparing recent cohorts to earlier ones reveals whether entrepreneurial ecosystem strength is improving or declining over time.
::: -->

## Startup-Scale vs. Qualified Establishments

The composition of new establishments by size reveals the nature of entrepreneurial activity. Startup-scale businesses (1-20 employees) represent **grassroots entrepreneurship**, while larger qualified establishments (21+ employees) often indicate **expansion activity** or **branch locations**.

::: column-page
```{r survival_by_size, fig.height=14}
#| fig-cap: "Survival rates and growth trajectories differ between startup-scale and larger qualified establishments. This analysis reveals which establishment types demonstrate greater resilience and growth potential across initiatives."

knitr::include_url("estab_output_2022/plots/H_survival_by_size_interactive.html", height = "1000px")
```
:::

## Sector-Level Entrepreneurship Patterns

Within each initiative, specific sectors demonstrate varying levels of entrepreneurial activity and business formation success.

```{r sector_entrepreneurship_table}
#| tbl-cap: "Top Entrepreneurial Sectors by Initiative"

sector_data <- read_csv("estab_output_2022/tables/R_sector_entrepreneurship_all_initiatives.csv", 
                        show_col_types = FALSE)

# Get top 5 sectors per initiative
sector_display <- sector_data %>%
  group_by(across(matches("initiative", ignore.case = TRUE))) %>%
  arrange(desc(across(matches("new.*establishments|formation.*count", ignore.case = TRUE)))) %>%
  slice_head(n = 5) %>%
  ungroup() %>%
  mutate(
    # Format year columns (no commas)
    across(matches("year", ignore.case = TRUE), ~as.character(.x)),
    # Format numeric columns (with commas)
    across(where(is.numeric) & !matches("year|rate|growth|pct", ignore.case = TRUE), ~comma(round(.x, 1))),
    # Format rate/percentage columns (just add %)
    across(matches("rate|pct", ignore.case = TRUE) & where(is.numeric), ~sprintf("%.1f%%", .x))
  ) %>%
  # Clean up column names
  rename_with(~str_replace_all(.x, "_", " ") %>% 
               str_to_title() %>%
               str_replace("Avg", "Average"))

create_table(sector_display, striped = TRUE)
```

::: {.callout-important icon="false"}
## 🎯 Sector-Specific Strategies

**Sector-level entrepreneurship patterns** inform targeted interventions:

- **High-formation sectors** with **strong survival rates** are proven areas for continued entrepreneurship support and cluster development
- **High-formation but low-survival sectors** may need enhanced business support services, mentorship programs, or market access assistance
- **Low-formation sectors** with **high survival rates** suggest high barriers to entry but sustainable business models once established
- **Concentration of formation** in specific sectors indicates where entrepreneurial momentum exists and where it may need cultivation

Understanding which sectors drive new business creation within each initiative helps **allocate entrepreneurship resources effectively** and **identify emerging specializations** before they appear in employment data.
:::

## New Startup Formation by Type

Breaking down startup formation by establishment type (true startups vs. qualified establishments) provides granular insights into entrepreneurial dynamics.

::: column-page
```{r startup_by_type, fig.height=14}
#| fig-cap: "Distinguishing between true startups and qualified establishments reveals whether growth is driven by grassroots entrepreneurship or larger-scale business development. Use dropdowns to explore different geographies and time periods."

knitr::include_url("estab_output_2022/plots/O_new_startups_by_type_overview.html", height = "1000px")
```
:::

::: column-page
```{r startup_by_type_subcluster, fig.height=14}
#| fig-cap: "Distinguishing between true startups and qualified establishments reveals whether growth is driven by grassroots entrepreneurship or larger-scale business development. Use dropdowns to explore different geographies, initiatives, subclusters, and time periods."

knitr::include_url("estab_output_2022/plots/O_new_startups_by_subcluster.html", height = "1000px")
```
:::

## Growth and Survival Patterns

Understanding how new establishments grow and survive over time distinguishes **sustainable business formation** from **churn**. Survival analysis reveals which initiatives and geographies foster **durable entrepreneurship**.

::: column-page
```{r growth_survival_bubble, fig.height=14}
#| fig-cap: "Bubble charts position establishments by employment growth and survival rates. Larger bubbles represent more establishments. High-growth, high-survival quadrants indicate robust entrepreneurial ecosystems."

knitr::include_url("estab_output_2022/plots/J_growth_survival_scale_bubble_interactive.html", height = "1000px")
```
:::

::: {.callout-important icon="false"}
## 🎯 Ecosystem Health Indicators

Strong entrepreneurial ecosystems demonstrate:

- **High formation rates** of new establishments
- **Strong survival rates** beyond initial years
- **Employment growth** in surviving establishments
- **Rising wage quality** as businesses mature
- **Balanced mix** of startup-scale and qualified establishments

Regions exhibiting these characteristics possess the **entrepreneurial infrastructure**, **talent pools**, and **market conditions** that support sustainable business creation. Understanding where these conditions exist—and where they're lacking—informs **targeted interventions** to strengthen entrepreneurial ecosystems across Indiana's advanced industries.
:::

# Talent Demand & Occupations {#sec-occupations}

Understanding the **occupational composition** of Indiana's advanced industries reveals the specific skills and talent required to support these sectors. This section examines the **most in-demand occupations**, their **distribution across metro areas**, and the critical role of **STEM and TECH occupations** in driving innovation and competitiveness.

## Top Occupations by Initiative

Each initiative requires distinct occupational profiles reflecting its unique workforce needs. Identifying the largest occupations within each initiative helps inform **talent development strategies** and **workforce training priorities**.

::: column-page
```{r top_occupations_viz, fig.height=14}
#| fig-cap: "The most in-demand occupations vary significantly across initiatives, reflecting distinct skill requirements and workforce compositions. Use the dropdown to examine different initiatives and geographies."

knitr::include_url(file.path(output_dir_occ, "visualizations/01_top_occupations_by_initiative.html"), height = "1200px")
```
:::

## Regional Occupation Patterns

Metro areas develop specialized talent pools aligned with their dominant industries. Understanding geographic variation in occupational demand reveals **regional workforce strengths** and **talent gaps**.

::: column-page
```{r metro_occupation_comparison, fig.height=14}
#| fig-cap: "Top occupations differ across metro areas within each initiative, revealing distinct regional talent specializations. Metro-level patterns inform targeted workforce development and talent recruitment strategies."

knitr::include_url(file.path(output_dir_occ, "visualizations/07_metro_comparison_top_occupations.html"), height = "1200px")
```
:::

## STEM Workforce Analysis

**STEM occupations** (Science, Technology, Engineering, and Mathematics) represent the innovation core of advanced industries. Tracking STEM employment trends reveals the state's capacity for **technological advancement** and **knowledge-intensive production**.

::: column-page
```{r stem_trends_viz, fig.height=12}
#| fig-cap: "STEM employment trends show the evolution of innovation-intensive occupations within each initiative. Solid lines represent STEM occupations while dashed lines show Non-STEM employment. Use the dropdown to examine different initiatives."

knitr::include_url(file.path(output_dir_occ, "visualizations/12a_stem_trends_interactive.html"), height = "900px")
```
:::
```{r stem_summary_table}
#| tbl-cap: "STEM Employment by Initiative and Geography"

stem_summary <- read_csv(file.path(output_dir_occ, "stem_employment_summary.csv"), 
                         show_col_types = FALSE)

# Calculate STEM percentage and get US + Indiana + top 3 metros per initiative
stem_display <- stem_summary %>%
  mutate(stem_pct = (stem_jobs / total_jobs) * 100) %>%
  # Identify US and Indiana
  mutate(
    is_us = geo_area == "United States",
    is_indiana = geo_area == "Indiana",
    is_benchmark = is_us | is_indiana
  ) %>%
  group_by(cicp_initiative) %>%
  arrange(desc(is_us), desc(is_indiana), desc(stem_pct)) %>%
  # Keep US + Indiana + top 3 metros
  filter(is_benchmark | (row_number() - sum(is_benchmark) <= 3)) %>%
  slice_head(n = 5) %>%  # US + IN + 3 metros = 5 total
  ungroup() %>%
  arrange(cicp_initiative, desc(is_us), desc(is_indiana), desc(stem_pct)) %>%
  mutate(
    `Total Jobs` = comma(round(total_jobs, 0)),
    `STEM Jobs` = comma(round(stem_jobs, 0)),
    `Non-STEM Jobs` = comma(round(nonstem_jobs, 0)),
    `STEM %` = sprintf("%.1f%%", stem_pct)
  ) %>%
  select(
    Initiative = cicp_initiative,
    Geography = geo_area,
    `Total Jobs`,
    `STEM Jobs`,
    `Non-STEM Jobs`,
    `STEM %`
  )

create_table(stem_display, striped = TRUE)
```

::: {.callout-note icon="false"}
## 💡 STEM Workforce Insights

**STEM concentration** varies substantially across initiatives and geographies:

- Higher STEM percentages indicate **knowledge-intensive** production and **innovation capacity**
- Geographic variation reveals where **technical talent** is concentrated
- Trends over time show whether initiatives are becoming **more technology-driven**

:::

## TECH Workforce Analysis

**TECH occupations** represent the subset of technical roles directly involved in **computing, data, and digital technologies**. These occupations are increasingly critical across all advanced industries as **digitalization** and **automation** accelerate.

::: column-page
```{r tech_trends_viz, fig.height=12}
#| fig-cap: "TECH employment trends reveal the penetration of digital and computing occupations across initiatives. Solid lines represent TECH occupations while dashed lines show Non-TECH employment. Use the dropdown to examine different initiatives."

knitr::include_url(file.path(output_dir_occ, "visualizations/12b_tech_trends_interactive.html"), height = "900px")
```
:::
```{r tech_summary_table}
#| tbl-cap: "TECH Employment by Initiative and Geography"

tech_summary <- read_csv(file.path(output_dir_occ, "tech_employment_summary.csv"), 
                         show_col_types = FALSE)

# Calculate TECH percentage and get US + Indiana + top 3 metros per initiative
tech_display <- tech_summary %>%
  mutate(tech_pct = (tech_jobs / total_jobs) * 100) %>%
  # Identify US and Indiana
  mutate(
    is_us = geo_area == "United States",
    is_indiana = geo_area == "Indiana",
    is_benchmark = is_us | is_indiana
  ) %>%
  group_by(cicp_initiative) %>%
  arrange(desc(is_us), desc(is_indiana), desc(tech_pct)) %>%
  # Keep US + Indiana + top 3 metros
  filter(is_benchmark | (row_number() - sum(is_benchmark) <= 3)) %>%
  slice_head(n = 5) %>%  # US + IN + 3 metros = 5 total
  ungroup() %>%
  arrange(cicp_initiative, desc(is_us), desc(is_indiana), desc(tech_pct)) %>%
  mutate(
    `Total Jobs` = comma(round(total_jobs, 0)),
    `TECH Jobs` = comma(round(tech_jobs, 0)),
    `Non-TECH Jobs` = comma(round(nontech_jobs, 0)),
    `TECH %` = sprintf("%.1f%%", tech_pct)
  ) %>%
  select(
    Initiative = cicp_initiative,
    Geography = geo_area,
    `Total Jobs`,
    `TECH Jobs`,
    `Non-TECH Jobs`,
    `TECH %`
  )

create_table(tech_display, striped = TRUE)
```

::: {.callout-note icon="false"}
## 💡 TECH Workforce Insights

**TECH occupation concentration** reflects digital transformation progress:

- TECH roles are expanding beyond traditional technology sectors into all advanced industries
- Geographic patterns reveal **technology talent hubs** and potential **skills gaps**

:::

# Methodology & Technical Notes {#sec-methodology}

## Geographic Coverage

Analysis includes three geographic levels:

-   **United States**: National benchmarks for comparative analysis
-   **Indiana**: Statewide aggregates
-   **Metro Areas**: Metropolitan Statistical Areas (MSAs) within Indiana

## Key Metrics Definitions

```{r methodology_metrics}
#| tbl-cap: "Metric Definitions and Interpretations"

metrics <- tribble(
  ~Metric, ~Definition, ~Interpretation,
  "Jobs", "Total employment in sector/geography", "Scale measure",
  "Wages", "Average annual wages per worker", "Economic value per worker",
  "GDP/GRP", "Gross Regional Product", "Total economic output",
  "YoY Growth", "Year-over-year % change", "Short-term momentum",
  "CAGR", "Compound Annual Growth Rate", "Sustained trend",
  "GDP per Job", "GDP ÷ Employment", "Productivity measure",
  "Location Quotient", "(Local %) ÷ (National %)", "Specialization measure (>1 = specialized)"
)

create_table(metrics, striped = TRUE)
```

## Analytical Approaches

### Growth Calculations

-   **Year-over-Year (YoY)**: $(Value_t / Value_{t-1}) - 1$
-   **CAGR**: $(Value_{end} / Value_{start})^{1/years} - 1$
-   **Productivity Change**: GDP Growth Rate - Jobs Growth Rate

### Location Quotient

$LQ = \frac{(Regional\ Jobs_i / Regional\ Jobs_{total})}{(National\ Jobs_i / National\ Jobs_{total})}$

Where: - $LQ > 1$: Regional specialization (higher concentration than national average) - $LQ = 1$: Regional concentration matches national average - $LQ < 1$: Regional underrepresentation

### Quadrant Analysis

Uses **median values** as thresholds to classify metros: - **Stars**: Above-median growth AND above-median employment share - **Emerging**: Above-median growth, below-median employment share - **Mature**: Below-median growth, above-median employment share - **Declining**: Below-median growth AND below-median employment share

## Data Quality Notes

-   Employment data sourced from **Quarterly Census of Employment and Wages (QCEW) via Lightcast**
-   GDP/GRP data from **Bureau of Economic Analysis (BEA)**
-   Some metro areas may have **suppressed values** for confidentiality
-   **Projections** (years beyond 2024) indicated by dashed lines in visualizations
-   All monetary values in **nominal dollars** (not inflation-adjusted)

## Limitations & Caveats

::: callout-warning
## ⚠️ Important Considerations

1.  **Initiative Overlap**: Some jobs may be counted in multiple initiatives if they meet criteria for more than one classification
2.  **Geographic Boundaries**: Metro area definitions follow OMB standards and include the *entire* metros, not just the Indiana portions
3.  **Timing Lags**: GDP data may lag employment data by several months
4.  **Suppressed Data**: Small cell sizes suppressed for confidentiality may affect some calculations
5.  **Projection Uncertainty**: Future year estimates (beyond 2024) carry increasing uncertainty
6.  **Seasonal Adjustment**: Some metrics are not seasonally adjusted
:::

# Additional Resources {#sec-resources}

## Data Sources

-   **CICP Advanced Industries Dashboard**: <https://www.cicpindiana.com/advanced-industry-dashboard/>
-   **Bureau of Labor Statistics QCEW**: <https://www.bls.gov/cew/>
-   **Bureau of Economic Analysis**: <https://www.bea.gov/>
-   **Lightcast**: <https://www.lightcast.io/>

## Initiative Websites

-   **CICP**: <https://www.cicpindiana.com/>
-   **AgriNovus Indiana**: <https://agrinovusindiana.com/>
-   **BioCrossroads**: <https://biocrossroads.com/>
-   **Conexus Indiana**: <https://conexusindiana.com/>
-   **TechPoint**: <https://techpoint.org/>

## Contact Information

For questions, additional analysis, or data requests:

**Riley Hudelson-Zipper**\
Email: zipperr\@iu.edu

------------------------------------------------------------------------

::: {.callout-note icon="false" appearance="minimal"}
## 📄 Report Information

**Report Generated**: `r format(Sys.Date(), "%B %d, %Y")`\
**Data Period**: `r min(jobs_data$year)` – `r max(jobs_data$year)`\
**Most Recent Complete Year**: `r recent_year`\
**Analysis Version**: v2.0\
**Output Directory**: `r output_dir`

------------------------------------------------------------------------

*This report was generated using R, Quarto, and the tidyverse ecosystem. Interactive visualizations created with Plotly.*
:::

\`\`\`