---
title: "CICP Supplemental Report"
date: today
date-format: "MMMM D, YYYY"
author: "Riley Hudelson-Zipper"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    theme: 
      - cosmo
      - custom.scss
    fontsize: 1.1em
    fig-width: 18
    fig-height: 9
    fig-align: center
    embed-resources: true
    smooth-scroll: true
    link-external-newwindow: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false

library(tidyverse)
library(scales)
library(knitr)
library(kableExtra)
library(plotly)
library(htmlwidgets)
library(DT)
library(janitor)
library(readxl)

# Helper function for beautiful tables
create_table <- function(df, title = NULL, striped = TRUE) {
  df %>%
    kable(
      format = "html",
      caption = title,
      digits = 1,
      format.args = list(big.mark = ","),
      align = "l",
      escape = FALSE
    ) %>%
    kable_styling(
      bootstrap_options = c(
        if(striped) "striped" else NULL, 
        "hover", 
        "condensed", 
        "responsive"
      ),
      full_width = FALSE,
      position = "left",
      font_size = 13
    ) %>%
    row_spec(0, bold = TRUE, background = "#f8f9fa") %>%
    as.character() %>%
    htmltools::HTML()
}

```

# Introduction

This supplemental report examines **cross-cutting economic metrics** that provide important context for understanding Indiana's advanced industry ecosystem but don't fit neatly within the initiative- or industry-specific analyses presented in the main report.

While the primary CICP report focuses on employment, wages, and GDP trends within specific advanced industry initiatives (TechPoint, BioCrossroads, Conexus, AgriNovus, and the broader Advanced & Traded Industries category), this analysis explores **economy-wide performance indicators** that influence and reflect the competitiveness of these sectors.

## Scope of Analysis

This report includes:

- **Innovation Metrics**: Research and development investment, patent activity, and other indicators of Indiana's innovation capacity (covered in separate R&D analysis)
- **Labor Productivity Trends**: Bureau of Labor Statistics data on output per hour worked, providing insights into efficiency gains and competitiveness across industries
- **Additional Economic Indicators**: As relevant data becomes available, this report may expand to include other cross-cutting metrics that inform advanced industry policy and strategy

These metrics complement rather than duplicate the main report's industry-focused analysis, offering a broader economic context for evaluating Indiana's advanced industry performance relative to national trends and peer states.

# State R&D Performance {#sec-rd-performance}

Indiana's research and development activity is a critical driver of innovation and advanced industry competitiveness. This section examines R&D value added, employment, and compensation using data from the [Bureau of Economic Analysis R&D Satellite Account](https://www.bea.gov/data/special-topics/research-and-development-satellite-account).

```{r rd_data_import}
#| include: false

# Import R&D data
rd_raw <- read_excel(file.path("CICP_20250709", "rd-satellite-account-va-emp-comp-2012-2023.xlsx"))

# Separate notes (rows starting from 2343)
rd_notes <- rd_raw %>%
  slice(2343:n())

# Clean main data
rd_data <- rd_raw %>%
  slice(1:2342) %>%
  clean_names() %>%
  # Remove superscripts from description
  mutate(description = str_remove_all(description, "\\d+$")) %>%
  # Filter for US, Indiana, and neighboring states
  filter(geo_fips %in% c("00000", "18000", "26000", "17000", "39000", "21000")) %>%
  # Convert to long format
  pivot_longer(
    cols = matches("^x?\\d{4}$"),
    names_to = "year",
    values_to = "value"
  ) %>%
  # Clean year column (remove 'x' prefix if present)
  mutate(
    year = as.numeric(str_remove(year, "^x")),
    # Convert (L) to NA
    value = if_else(value == "(L)", NA_character_, value),
    value = as.numeric(value)
  ) %>%
  # Create cleaner geography name
  mutate(
    geography = case_when(
      geo_fips == "00000" ~ "United States",
      geo_fips == "18000" ~ "Indiana",
      geo_fips == "26000" ~ "Michigan",
      geo_fips == "17000" ~ "Illinois",
      geo_fips == "39000" ~ "Ohio",
      geo_fips == "21000" ~ "Kentucky",
      TRUE ~ geo_name
    )
  )

# Create separate datasets by table type
rd_value_added <- rd_data %>% filter(table_name == "SARDVA")
rd_employment <- rd_data %>% filter(table_name == "SARDEMP")
rd_compensation <- rd_data %>% filter(table_name == "SARDCOMP")

# Define color palette for states
state_colors <- c(
  "Indiana" = "#1565C0",
  "Michigan" = "#D84315",
  "Illinois" = "#6A1B9A",
  "Ohio" = "#2E7D32",
  "Kentucky" = "#F57C00",
  "United States" = "#424242"
)
```

## R&D Value Added Trends

R&D value added measures the economic contribution of research and development activities.

```{r rd_value_trends}
#| fig-cap: "R&D Value Added: Indiana and Neighboring States vs. United States (Indexed to 2012)"
#| fig-height: 16

# Filter for top-level R&D and index to 2012
rd_va_total <- rd_value_added %>%
  filter(line_code == 1) %>%  # Total R&D
  select(geography, year, value, description, unit) %>%
  group_by(geography) %>%
  mutate(
    base_value = first(value[year == 2012]),
    indexed_value = (value / base_value) * 100
  ) %>%
  ungroup()

# Create interactive line chart
p_rd_va <- plot_ly()

# 1. Add US first (dashed reference line)
us_data <- rd_va_total %>% filter(geography == "United States")
p_rd_va <- p_rd_va %>%
  add_trace(
    data = us_data,
    x = ~year, y = ~indexed_value,
    name = "United States",
    type = 'scatter', mode = 'lines',
    line = list(color = state_colors["United States"], width = 2, dash = "dash"),
    hovertemplate = paste0(
      "<b>United States</b><br>",
      "Year: %{x}<br>",
      "Indexed Value: %{y:.1f}<br>",
      "Actual Value: $%{customdata:,.0f}M<br>",
      "<extra></extra>"
    ),
    customdata = ~value
  )

# 2. Add Indiana second (thicker line)
indiana_data <- rd_va_total %>% filter(geography == "Indiana")
p_rd_va <- p_rd_va %>%
  add_trace(
    data = indiana_data,
    x = ~year, y = ~indexed_value,
    name = "Indiana",
    type = 'scatter', mode = 'lines+markers',
    line = list(color = state_colors["Indiana"], width = 5),  # Thicker line
    marker = list(color = state_colors["Indiana"], size = 10),  # Larger markers too
    hovertemplate = paste0(
      "<b>Indiana</b><br>",
      "Year: %{x}<br>",
      "Indexed Value: %{y:.1f}<br>",
      "Actual Value: $%{customdata:,.0f}M<br>",
      "<extra></extra>"
    ),
    customdata = ~value
  )

# 3. Add other states (in alphabetical order)
other_states <- rd_va_total %>% 
  filter(!geography %in% c("United States", "Indiana")) %>%
  arrange(geography) %>%
  distinct(geography) %>%
  pull(geography)

for(state in other_states) {
  state_subset <- rd_va_total %>% filter(geography == state)
  p_rd_va <- p_rd_va %>%
    add_trace(
      data = state_subset,
      x = ~year, y = ~indexed_value,
      name = state,
      type = 'scatter', mode = 'lines+markers',
      line = list(color = state_colors[state], width = 3),
      marker = list(color = state_colors[state], size = 8),
      hovertemplate = paste0(
        "<b>", state, "</b><br>",
        "Year: %{x}<br>",
        "Indexed Value: %{y:.1f}<br>",
        "Actual Value: $%{customdata:,.0f}M<br>",
        "<extra></extra>"
      ),
      customdata = ~value
    )
}

p_rd_va <- p_rd_va %>%
  layout(
    title = list(
      text = "<b>Total R&D Value Added by State (Indexed to 2012 = 100)</b><br><sup>Easier comparison of relative growth trajectories</sup>",
      font = list(size = 16)
    ),
    xaxis = list(title = "Year"),
    yaxis = list(title = "Indexed Value (2012 = 100)"),
    hovermode = "x unified",
    legend = list(
      orientation = "v", 
      x = 1.02, 
      y = 0.5,
      traceorder = "normal"  # Keeps the order traces were added
    ),
    shapes = list(
      list(
        type = "line",
        x0 = 2012,
        x1 = 2023,
        y0 = 100,
        y1 = 100,
        line = list(color = "gray", width = 1, dash = "dot")
      )
    )
  )

p_rd_va
```

```{r rd_va_summary_table}
#| tbl-cap: "R&D Value Added Summary Statistics by State (2012-2023)"

rd_va_summary <- rd_va_total %>%
  group_by(geography) %>%
  summarise(
    `2012` = dollar(first(value[year == 2012]), scale = 1e-3, suffix = "M", accuracy = .1),
    `2023` = dollar(last(value[year == 2023]), scale = 1e-3, suffix = "M", accuracy = .1),
    `Total Growth` = sprintf("%.1f%%", ((last(value) / first(value)) - 1) * 100),
    `CAGR` = sprintf("%.1f%%", ((last(value) / first(value))^(1/11) - 1) * 100),
    cagr_numeric = ((last(value) / first(value))^(1/11) - 1) * 100,  # Keep numeric for sorting
    .groups = "drop"
  ) %>%
  # Create custom sort order
  mutate(
    sort_order = case_when(
      geography == "United States" ~ 1,
      geography == "Indiana" ~ 2,
      TRUE ~ 3
    )
  ) %>%
  arrange(sort_order, desc(cagr_numeric)) %>%
  select(-sort_order, -cagr_numeric)

create_table(rd_va_summary)
```

## Regional R&D Comparison

How does Indiana's R&D intensity compare to neighboring states?
```{r rd_state_comparison}
#| fig-cap: "R&D Value Added by State (2023)"
#| fig-height: 8

# 2023 comparison
rd_2023_comparison <- rd_va_total %>%
  filter(geography != "United States", year == 2023) %>%
  arrange(desc(value))

p_rd_comparison <- plot_ly(
  rd_2023_comparison,
  x = ~value,
  y = ~reorder(geography, value),
  type = 'bar',
  orientation = 'h',
  marker = list(
    color = ~geography,
    colors = state_colors
  ),
  hovertemplate = paste0(
    "<b>%{y}</b><br>",
    "Value Added: $%{x:,.1f}M<br>",
    "<extra></extra>"
  )
) %>%
  layout(
    title = list(
      text = "<b>R&D Value Added by State (2023)</b><br><sup>Millions of Current Dollars</sup>",
      font = list(size = 16)
    ),
    xaxis = list(title = "Value Added ($ Millions)", tickformat = ","),
    yaxis = list(title = ""),
    showlegend = FALSE,
    margin = list(l = 100)
  )

p_rd_comparison
```

## R&D Employment Comparison

R&D employment captures the workforce dedicated to research and development activities across the region.
```{r rd_employment_trends}
#| fig-cap: "R&D Employment: Indiana and Neighboring States (Indexed to 2012)"
#| fig-height: 14

# Filter for top-level R&D employment and index to 2012
rd_emp_total <- rd_employment %>%
  filter(line_code == 1) %>%
  select(geography, year, value, description, unit) %>%
  group_by(geography) %>%
  mutate(
    base_value = first(value[year == 2012]),
    indexed_value = (value / base_value) * 100
  ) %>%
  ungroup()

# Create interactive line chart
p_rd_emp <- plot_ly()

# 1. Add US first (dashed reference line)
us_emp_data <- rd_emp_total %>% filter(geography == "United States")
p_rd_emp <- p_rd_emp %>%
  add_trace(
    data = us_emp_data,
    x = ~year, y = ~indexed_value,
    name = "United States",
    type = 'scatter', mode = 'lines',
    line = list(color = state_colors["United States"], width = 2, dash = "dash"),
    hovertemplate = paste0(
      "<b>United States</b><br>",
      "Year: %{x}<br>",
      "Indexed Value: %{y:.1f}<br>",
      "Actual Employment (thousands of jobs): %{customdata:,}<br>",
      "<extra></extra>"
    ),
    customdata = ~value
  )

# 2. Add Indiana second (thicker line)
indiana_emp_data <- rd_emp_total %>% filter(geography == "Indiana")
p_rd_emp <- p_rd_emp %>%
  add_trace(
    data = indiana_emp_data,
    x = ~year, y = ~indexed_value,
    name = "Indiana",
    type = 'scatter', mode = 'lines+markers',
    line = list(color = state_colors["Indiana"], width = 5),
    marker = list(color = state_colors["Indiana"], size = 10),
    hovertemplate = paste0(
      "<b>Indiana</b><br>",
      "Year: %{x}<br>",
      "Indexed Value: %{y:.1f}<br>",
      "Actual Employment (thousands of jobs): %{customdata:,}<br>",
      "<extra></extra>"
    ),
    customdata = ~value
  )

# 3. Add other states (in alphabetical order)
other_states_emp <- rd_emp_total %>% 
  filter(!geography %in% c("United States", "Indiana")) %>%
  arrange(geography) %>%
  distinct(geography) %>%
  pull(geography)

for(state in other_states_emp) {
  state_subset <- rd_emp_total %>% filter(geography == state)
  p_rd_emp <- p_rd_emp %>%
    add_trace(
      data = state_subset,
      x = ~year, y = ~indexed_value,
      name = state,
      type = 'scatter', mode = 'lines+markers',
      line = list(color = state_colors[state], width = 3),
      marker = list(color = state_colors[state], size = 8),
      hovertemplate = paste0(
        "<b>", state, "</b><br>",
        "Year: %{x}<br>",
        "Indexed Value: %{y:.1f}<br>",
        "Actual Employment (thousands of jobs): %{customdata:,}<br>",
        "<extra></extra>"
      ),
      customdata = ~value
    )
}

p_rd_emp <- p_rd_emp %>%
  layout(
    title = list(
      text = "<b>Total R&D Employment by State (Indexed to 2012 = 100)</b><br><sup>Easier comparison of relative growth trajectories</sup>",
      font = list(size = 16)
    ),
    xaxis = list(title = "Year"),
    yaxis = list(title = "Indexed Employment (2012 = 100)"),
    hovermode = "x unified",
    legend = list(
      orientation = "v", 
      x = 1.02, 
      y = 0.5,
      traceorder = "normal"
    ),
    shapes = list(
      list(
        type = "line",
        x0 = 2012,
        x1 = 2023,
        y0 = 100,
        y1 = 100,
        line = list(color = "gray", width = 1, dash = "dot")
      )
    )
  )

p_rd_emp
```
```{r rd_emp_summary_table}
#| tbl-cap: "R&D Employment (in Thousands of Jobs) Summary by State (2012-2023)"

rd_emp_summary <- rd_emp_total %>%
  group_by(geography) %>%
  summarise(
    `2012` = comma(first(value[year == 2012])),
    `2023` = comma(last(value[year == 2023])),
    `Total Growth` = sprintf("%.1f%%", ((last(value) / first(value)) - 1) * 100),
    `CAGR` = sprintf("%.1f%%", ((last(value) / first(value))^(1/11) - 1) * 100),
    cagr_numeric = ((last(value) / first(value))^(1/11) - 1) * 100,
    .groups = "drop"
  ) %>%
  mutate(
    sort_order = case_when(
      geography == "United States" ~ 1,
      geography == "Indiana" ~ 2,
      TRUE ~ 3
    )
  ) %>%
  arrange(sort_order, desc(cagr_numeric)) %>%
  select(-sort_order, -cagr_numeric)

create_table(rd_emp_summary)
```

## R&D Compensation per Worker

Examining compensation per R&D worker reveals the economic value and competitiveness of R&D employment across states.

```{r rd_compensation_per_worker}
#| tbl-cap: "R&D Compensation per Worker by State (2023)"

# Calculate compensation per worker
rd_comp_per_worker <- rd_compensation %>%
  filter(line_code == 1) %>%
  left_join(
    rd_employment %>% filter(line_code == 1) %>% select(geography, year, employment = value),
    by = c("geography", "year")
  ) %>%
  mutate(comp_per_worker = (value * 1000000) / (employment * 1000)) %>%
  filter(year == 2023) %>%
  arrange(desc(comp_per_worker)) %>%
  select(geography, `Total Compensation` = value, Employment = employment, comp_per_worker) %>%
  mutate(
    `Total Compensation` = dollar(`Total Compensation`, suffix = "M", accuracy = .1),
    Employment = comma(Employment, suffix = "K", accuracy = 1),
    `Compensation per Worker` = dollar(comp_per_worker, accuracy = 1)
  ) %>%
  select(State = geography, `Total Compensation`, Employment, `Compensation per Worker`)

create_table(rd_comp_per_worker)
```
```{r rd_comp_comparison_viz}
#| fig-cap: "R&D Compensation per Worker by State (2023)"
#| fig-height: 8

# Visualization of compensation per worker
rd_comp_viz_data <- rd_compensation %>%
  filter(line_code == 1) %>%
  left_join(
    rd_employment %>% filter(line_code == 1) %>% select(geography, year, employment = value),
    by = c("geography", "year")
  ) %>%
  mutate(comp_per_worker = (value * 1000000) / (employment * 1000)) %>%
  filter(year == 2023) %>%
  arrange(desc(comp_per_worker))

p_comp <- plot_ly(
  rd_comp_viz_data,
  x = ~comp_per_worker,
  y = ~reorder(geography, comp_per_worker),
  type = 'bar',
  orientation = 'h',
  marker = list(
    color = ~geography,
    colors = state_colors
  ),
  text = "",  # Remove text labels
  textposition = "none",
  hovertemplate = paste0(
    "<b>%{y}</b><br>",
    "Comp per Worker: $%{x:,.0f}<br>",
    "<extra></extra>"
  )
) %>%
  layout(
    title = list(
      text = "<b>R&D Compensation per Worker by State (2023)</b><br><sup>Average Annual Compensation</sup>",
      font = list(size = 16)
    ),
    xaxis = list(title = "Compensation per Worker ($)", tickformat = "$,"),
    yaxis = list(title = ""),
    showlegend = FALSE,
    margin = list(l = 100)
  )

p_comp
```

## R&D by Sector

Breaking down R&D activity by manufacturing vs. nonmanufacturing sectors reveals concentration patterns.

```{r rd_sector_time_series}
#| fig-cap: "R&D Value Added by Sector Over Time - All States"
#| fig-height: 16

# Get manufacturing and nonmanufacturing data for all geographies
rd_sector_time <- rd_value_added %>%
  filter(line_code %in% c(4, 9)) %>%  # 4 = Manufacturing, 9 = Nonmanufacturing
  select(geography, year, description, value)

# Create faceted interactive chart
p_rd_sector_time <- plot_ly()

# Get unique geographies
geographies <- c("United States", "Indiana", sort(setdiff(unique(rd_sector_time$geography), c("United States", "Indiana"))))

# Create subplot data
subplot_list <- list()

for(i in seq_along(geographies)) {
  geo <- geographies[i]
  geo_data <- rd_sector_time %>% filter(geography == geo)
  
  # Create subplot for this geography
  p_subplot <- plot_ly(geo_data) %>%
    add_trace(
      x = ~year, 
      y = ~value,
      color = ~description,
      colors = c("Manufacturing" = "#1565C0", "Nonmanufacturing" = "#2E7D32"),
      type = 'scatter',
      mode = 'lines+markers',
      line = list(width = 3),
      marker = list(size = 6),
      hovertemplate = paste0(
        "<b>%{fullData.name}</b><br>",
        "Year: %{x}<br>",
        "Value: $%{y:,.1f}M<br>",
        "<extra></extra>"
      ),
      showlegend = if(i == 1) TRUE else FALSE  # Only show legend for first subplot
    ) %>%
    layout(
      xaxis = list(title = if(i > length(geographies) - 2) "Year" else ""),
      yaxis = list(title = "Value Added ($M)", tickformat = ","),
      annotations = list(
        list(
          text = paste0("<b>", geo, "</b>"),
          x = 0.5,
          y = 1,
          xref = "paper",
          yref = "paper",
          xanchor = "center",
          yanchor = "bottom",
          showarrow = FALSE,
          font = list(size = 12)
        )
      )
    )
  
  subplot_list[[i]] <- p_subplot
}

# Combine into subplots
p_rd_sector_time <- subplot(
  subplot_list,
  nrows = 3,
  shareX = TRUE,
  titleY = TRUE,
  margin = 0.05
) %>%
  layout(
    title = list(
      text = "<b>R&D Value Added by Sector Over Time</b><br><sup>Manufacturing vs. Nonmanufacturing | Millions of Current Dollars</sup>",
      font = list(size = 16)
    ),
    showlegend = TRUE,
    legend = list(
      orientation = "h",
      x = 0.5,
      y = -0.05,
      xanchor = "center",
      yanchor = "top"
    ),
    margin = list(t = 150, l = 80)
  )

p_rd_sector_time
```


```{r rd_sector_breakdown}
#| fig-cap: "R&D Value Added by Sector - Indiana"
#| fig-height: 8

# Get sector breakdown for Indiana
rd_sector <- rd_value_added %>%
  filter(
    geography == "Indiana",
    line_code %in% c(4, 9),  # Manufacturing and Nonmanufacturing
    year == 2023
  ) %>%
  arrange(desc(value))

# Create bar chart
p_rd_sector <- plot_ly(
  rd_sector,
  x = ~value,
  y = ~reorder(description, value),
  type = 'bar',
  orientation = 'h',
  marker = list(color = c("#1565C0", "#2E7D32")),
  hovertemplate = paste0(
    "<b>%{y}</b><br>",
    "Value Added: $%{x:.1f}M<br>",
    "<extra></extra>"
  )
) %>%
  layout(
    title = list(
      text = "<b>R&D Value Added by Sector - Indiana (2023)</b><br><sup>Millions of Current Dollars</sup>",
      font = list(size = 16)
    ),
    xaxis = list(title = "Value Added ($ Millions)", tickformat = ","),
    yaxis = list(title = ""),
    margin = list(l = 150)
  )

p_rd_sector
```

::: {.callout-important icon="false"}
## ⚙️ Sector R&D VA Insights
```{r sector_insights}
#| output: asis

# Calculate key metrics
in_mfg_2023 <- rd_sector_time %>% filter(geography == "Indiana", description == "Manufacturing", year == 2023) %>% pull(value)
in_nonmfg_2023 <- rd_sector_time %>% filter(geography == "Indiana", description == "Nonmanufacturing", year == 2023) %>% pull(value)
in_mfg_share <- (in_mfg_2023 / (in_mfg_2023 + in_nonmfg_2023)) * 100

oh_mfg_2023 <- rd_sector_time %>% filter(geography == "Ohio", description == "Manufacturing", year == 2023) %>% pull(value)
oh_nonmfg_2023 <- rd_sector_time %>% filter(geography == "Ohio", description == "Nonmanufacturing", year == 2023) %>% pull(value)
oh_mfg_share <- (oh_mfg_2023 / (oh_mfg_2023 + oh_nonmfg_2023)) * 100

mi_mfg_2023 <- rd_sector_time %>% filter(geography == "Michigan", description == "Manufacturing", year == 2023) %>% pull(value)
mi_nonmfg_2023 <- rd_sector_time %>% filter(geography == "Michigan", description == "Nonmanufacturing", year == 2023) %>% pull(value)
mi_mfg_share <- (mi_mfg_2023 / (mi_mfg_2023 + mi_nonmfg_2023)) * 100

il_mfg_2023 <- rd_sector_time %>% filter(geography == "Illinois", description == "Manufacturing", year == 2023) %>% pull(value)
il_nonmfg_2023 <- rd_sector_time %>% filter(geography == "Illinois", description == "Nonmanufacturing", year == 2023) %>% pull(value)
il_mfg_share <- (il_mfg_2023 / (il_mfg_2023 + il_nonmfg_2023)) * 100

# Growth rates
in_mfg_growth <- rd_sector_time %>% 
  filter(geography == "Indiana", description == "Manufacturing", year %in% c(2012, 2023)) %>%
  summarise(growth = ((last(value) / first(value)) - 1) * 100) %>% pull(growth)

in_nonmfg_growth <- rd_sector_time %>% 
  filter(geography == "Indiana", description == "Nonmanufacturing", year %in% c(2012, 2023)) %>%
  summarise(growth = ((last(value) / first(value)) - 1) * 100) %>% pull(growth)

oh_mfg_growth <- rd_sector_time %>% 
  filter(geography == "Ohio", description == "Manufacturing", year %in% c(2012, 2023)) %>%
  summarise(growth = ((last(value) / first(value)) - 1) * 100) %>% pull(growth)

oh_nonmfg_growth <- rd_sector_time %>% 
  filter(geography == "Ohio", description == "Nonmanufacturing", year %in% c(2012, 2023)) %>%
  summarise(growth = ((last(value) / first(value)) - 1) * 100) %>% pull(growth)

cat("**Indiana's Manufacturing R&D Dominance:**\n\n")
cat(sprintf("- Manufacturing accounts for **%.0f%%** of Indiana's total R&D value added (vs. %.0f%% nonmanufacturing), reflecting the state's strong industrial base\n", in_mfg_share, 100 - in_mfg_share))
cat(sprintf("- Indiana's manufacturing R&D grew **%.1f%%** from 2012-2023, while nonmanufacturing R&D grew **%.1f%%**\n\n", in_mfg_growth, in_nonmfg_growth))

cat("**Regional Comparison:**\n\n")
cat(sprintf("- **Michigan** (%.0f%% mfg) and **Indiana** (%.0f%% mfg) show the highest manufacturing R&D concentration among neighbors\n", mi_mfg_share, in_mfg_share))
cat(sprintf("- **Illinois** has the most balanced R&D portfolio (%.0f%% mfg, %.0f%% nonmfg), reflecting its diversified economy with strong services sectors\n", il_mfg_share, 100 - il_mfg_share))
cat(sprintf("- **Ohio** presents a notable contrast: its nonmanufacturing R&D has **nearly converged** with manufacturing, driven by %.1f%% nonmanufacturing growth vs. %.1f%% manufacturing growth\n\n", oh_nonmfg_growth, oh_mfg_growth))

cat("**Strategic Implications for Indiana:**\n\n")
cat("- Ohio's convergence demonstrates that even manufacturing-intensive states can **diversify their R&D portfolios** through investments in technology services, healthcare innovation, and other knowledge-intensive sectors\n")
cat("- Indiana's **persistent and widening gap** suggests potential vulnerability to manufacturing sector disruptions and missed opportunities in high-growth nonmanufacturing R&D fields\n")
cat("- The state may benefit from **targeted initiatives** to strengthen R&D in emerging sectors like software, data analytics, healthcare technology, and professional services to create a more **resilient innovation ecosystem**\n")
```
:::

## Regional R&D Share Analysis

Understanding Indiana's position relative to the broader Midwest R&D ecosystem.
```{r rd_regional_share}
#| tbl-cap: "State Share of Regional R&D Activity (2023)"

# Calculate regional shares
regional_totals <- rd_va_total %>%
  filter(year == 2023, geography != "United States") %>%
  summarise(regional_total = sum(value, na.rm = TRUE))

rd_regional_share <- rd_va_total %>%
  filter(year == 2023, geography != "United States") %>%
  mutate(
    regional_total = regional_totals$regional_total,
    share_of_region = (value / regional_total) * 100
  ) %>%
  arrange(desc(share_of_region)) %>%
  select(geography, value, share_of_region) %>%
  mutate(
    `R&D Value Added` = dollar(value, suffix = "M", accuracy = 1),
    `Share of 5-State Region` = sprintf("%.1f%%", share_of_region)
  ) %>%
  select(State = geography, `R&D Value Added`, `Share of 5-State Region`)

create_table(rd_regional_share)
```

## R&D Value Added by Detailed Industry

Breaking down R&D by specific industries reveals concentration patterns and growth dynamics within the broader manufacturing and nonmanufacturing categories.
```{r rd_industry_detail_data}
#| include: false

# Get detailed industry data
rd_industry_detail <- rd_value_added %>%
  filter(
    line_code %in% c(5:8, 10:12, 16, 21),
    !is.na(value)
  ) %>%
  mutate(
    sector_group = case_when(
      line_code %in% 5:8 ~ "Manufacturing",
      line_code %in% 10:12 ~ "Nonmanufacturing",
      line_code == 16 ~ "Government",
      line_code == 21 ~ "Higher Education"
    )
  ) %>%
  select(geography, year, line_code, description, value, sector_group)
```

### Manufacturing R&D by Industry
```{r rd_mfg_industries_viz}
#| fig-cap: "Manufacturing R&D Value Added by Industry - Indiana and Neighboring States (2023)"
#| fig-height: 16

# 2023 manufacturing industries comparison
rd_mfg_2023 <- rd_industry_detail %>%
  filter(sector_group == "Manufacturing", year == 2023) %>%
  mutate(
    description_full = description,  # Keep original for hover
    description = case_when(
      str_detect(description, "Computer and electronic") ~ "Electronic product mfg.",
      str_detect(description, "Chemical") ~ "Chemical mfg.",
      str_detect(description, "Transportation equipment") ~ "Transportation equip. mfg.",
      str_detect(description, "Other manufacturing") ~ "Other",
      TRUE ~ description
    )
  ) %>%
  arrange(geography, desc(value))

# Create faceted bar chart
p_mfg_detail <- plot_ly()

geographies_mfg <- c("United States", "Indiana", sort(setdiff(unique(rd_mfg_2023$geography), c("United States", "Indiana"))))

subplot_list_mfg <- list()

for(i in seq_along(geographies_mfg)) {
  geo <- geographies_mfg[i]
  geo_data <- rd_mfg_2023 %>% filter(geography == geo)
  
  if(nrow(geo_data) > 0) {
    p_subplot <- plot_ly(
      geo_data,
      x = ~value,
      y = ~reorder(description, value),
      type = 'bar',
      orientation = 'h',
      marker = list(color = "#1565C0"),
      customdata = ~description_full,
      hovertemplate = paste0(
        "<b>%{customdata}</b><br>",
        "Value: $%{x:,.1f}M<br>",
        "<extra></extra>"
      ),
      showlegend = FALSE
    ) %>%
      layout(
        xaxis = list(title = ""),
        yaxis = list(title = ""),
        annotations = list(
          list(
            text = paste0("<b>", geo, "</b>"),
            x = 0.5,
            y = 1.08,
            xref = "paper",
            yref = "paper",
            xanchor = "center",
            yanchor = "bottom",
            showarrow = FALSE,
            font = list(size = 11)
          )
        ),
        margin = list(t = 40)
      )
    
    subplot_list_mfg[[i]] <- p_subplot
  }
}

p_mfg_detail <- subplot(
  subplot_list_mfg,
  nrows = 3,
  shareX = FALSE,
  titleY = TRUE,
  margin = 0.1
) %>%
  layout(
    title = list(
      text = "<b>Manufacturing R&D Value Added by Industry in (2023), Millions of Dollars</b><br><sup>Detailed industry breakdown</sup>",
      font = list(size = 16)
    ),
    margin = list(t = 150, l = 80)
  )

p_mfg_detail
```

```{r rd_mfg_industries_table}
#| tbl-cap: "Manufacturing R&D Industries - Indiana (2023)"

rd_mfg_in_table <- rd_industry_detail %>%
  filter(sector_group == "Manufacturing", geography == "Indiana", year == 2023) %>%
  arrange(desc(value)) %>%
  mutate(
    total_mfg = sum(value, na.rm = TRUE),
    share = (value / total_mfg) * 100,
    `R&D Value Added` = dollar(value, suffix = "M", accuracy = 0.1),
    `Share of Mfg R&D` = sprintf("%.1f%%", share)
  ) %>%
  select(Industry = description, `R&D Value Added`, `Share of Mfg R&D`)

create_table(rd_mfg_in_table)
```

### Nonmanufacturing R&D by Industry
```{r rd_nonmfg_industries_viz}
#| fig-cap: "Nonmanufacturing R&D Value Added by Industry - Indiana and Neighboring States (2023)"
#| fig-height: 16

# 2023 nonmanufacturing industries comparison
rd_nonmfg_2023 <- rd_industry_detail %>%
  filter(sector_group == "Nonmanufacturing", year == 2023) %>%
  mutate(
    description_full = description,  # Keep original for hover
    description = case_when(
      str_detect(description, "Professional") ~ "Professional svcs.",
      str_detect(description, "Information") ~ "Information",
      str_detect(description, "Other nonmanufacturing") ~ "Other",
      TRUE ~ description
    )
  ) %>%
  arrange(geography, desc(value))

# Create faceted bar chart
p_nonmfg_detail <- plot_ly()

geographies_nonmfg <- c("United States", "Indiana", sort(setdiff(unique(rd_nonmfg_2023$geography), c("United States", "Indiana"))))

subplot_list_nonmfg <- list()

for(i in seq_along(geographies_nonmfg)) {
  geo <- geographies_nonmfg[i]
  geo_data <- rd_nonmfg_2023 %>% filter(geography == geo)
  
  if(nrow(geo_data) > 0) {
    p_subplot <- plot_ly(
      geo_data,
      x = ~value,
      y = ~reorder(description, value),
      type = 'bar',
      orientation = 'h',
      marker = list(color = "#2E7D32"),
      customdata = ~description_full,
      hovertemplate = paste0(
        "<b>%{customdata}</b><br>",
        "Value: $%{x:,.1f}M<br>",
        "<extra></extra>"
      ),
      showlegend = FALSE
    ) %>%
      layout(
        xaxis = list(
          title = "",
          tickformat = ",.0f"
        ),
        yaxis = list(title = ""),
        annotations = list(
          list(
            text = paste0("<b>", geo, "</b>"),
            x = 0.5,
            y = 1.08,
            xref = "paper",
            yref = "paper",
            xanchor = "center",
            yanchor = "bottom",
            showarrow = FALSE,
            font = list(size = 11)
          )
        ),
        margin = list(t = 40)
      )
    
    subplot_list_nonmfg[[i]] <- p_subplot
  }
}

p_nonmfg_detail <- subplot(
  subplot_list_nonmfg,
  nrows = 3,
  shareX = FALSE,
  titleY = TRUE,
  margin = 0.1
) %>%
  layout(
    title = list(
      text = "<b>Nonmanufacturing R&D Value Added by Industry in (2023), Millions of Dollars</b><br><sup>Detailed industry breakdown</sup>",
      font = list(size = 16)
    ),
    margin = list(t = 150, l = 80)
  )

p_nonmfg_detail
```

```{r rd_nonmfg_industries_table}
#| tbl-cap: "Nonmanufacturing R&D Industries - Indiana (2023)"

rd_nonmfg_in_table <- rd_industry_detail %>%
  filter(sector_group == "Nonmanufacturing", geography == "Indiana", year == 2023) %>%
  arrange(desc(value)) %>%
  mutate(
    total_nonmfg = sum(value, na.rm = TRUE),
    share = (value / total_nonmfg) * 100,
    `R&D Value Added` = dollar(value, suffix = "M", accuracy = 0.1),
    `Share of Nonmfg R&D` = sprintf("%.1f%%", share)
  ) %>%
  select(Industry = description, `R&D Value Added`, `Share of Nonmfg R&D`)

create_table(rd_nonmfg_in_table)
```

### Government and Higher Education R&D

Government and university R&D represent critical components of the regional innovation ecosystem, supporting basic research and workforce development.
```{r rd_govt_university_viz}
#| fig-cap: "Government and Higher Education R&D Value Added Over Time (Indexed to 2012)"
#| fig-height: 16

# Get government and university R&D indexed to 2012
rd_govt_univ <- rd_industry_detail %>%
  filter(sector_group %in% c("Government", "Higher Education")) %>%
  group_by(geography, description) %>%
  mutate(
    base_value = first(value[year == 2012]),
    indexed_value = (value / base_value) * 100
  ) %>%
  ungroup()

# Create faceted chart by sector
sector_groups <- c("Government", "Higher Education")
subplot_list_gu <- list()

for(i in seq_along(sector_groups)) {
  sector <- sector_groups[i]
  sector_data <- rd_govt_univ %>% filter(sector_group == sector)
  
  # Create subplot for this sector
  p_subplot <- plot_ly()
  
  # 1. Add US first (dashed reference line)
  us_data <- sector_data %>% filter(geography == "United States")
  p_subplot <- p_subplot %>%
    add_trace(
      data = us_data,
      x = ~year, y = ~indexed_value,
      name = "United States",
      type = 'scatter', mode = 'lines',
      line = list(color = state_colors["United States"], width = 2, dash = "dash"),
      customdata = ~value,
      hovertemplate = paste0(
        "<b>United States</b><br>",
        "Year: %{x}<br>",
        "Indexed: %{y:.1f}<br>",
        "Actual: $%{customdata:,.1f}M<br>",
        "<extra></extra>"
      ),
      showlegend = if(i == 1) TRUE else FALSE
    )
  
  # 2. Add Indiana (thicker line)
  indiana_data <- sector_data %>% filter(geography == "Indiana")
  p_subplot <- p_subplot %>%
    add_trace(
      data = indiana_data,
      x = ~year, y = ~indexed_value,
      name = "Indiana",
      type = 'scatter', mode = 'lines+markers',
      line = list(color = state_colors["Indiana"], width = 5),
      marker = list(color = state_colors["Indiana"], size = 10),
      customdata = ~value,
      hovertemplate = paste0(
        "<b>Indiana</b><br>",
        "Year: %{x}<br>",
        "Indexed: %{y:.1f}<br>",
        "Actual: $%{customdata:,.1f}M<br>",
        "<extra></extra>"
      ),
      showlegend = if(i == 1) TRUE else FALSE
    )
  
  # 3. Add other states (alphabetically)
  other_states_gu <- sector_data %>% 
    filter(!geography %in% c("United States", "Indiana")) %>%
    arrange(geography) %>%
    distinct(geography) %>%
    pull(geography)
  
  for(state in other_states_gu) {
    state_subset <- sector_data %>% filter(geography == state)
    p_subplot <- p_subplot %>%
      add_trace(
        data = state_subset,
        x = ~year, y = ~indexed_value,
        name = state,
        type = 'scatter', mode = 'lines+markers',
        line = list(color = state_colors[state], width = 3),
        marker = list(color = state_colors[state], size = 8),
        customdata = ~value,
        hovertemplate = paste0(
          "<b>", state, "</b><br>",
          "Year: %{x}<br>",
          "Indexed: %{y:.1f}<br>",
          "Actual: $%{customdata:,.1f}M<br>",
          "<extra></extra>"
        ),
        showlegend = if(i == 1) TRUE else FALSE
      )
  }
  
  # Layout for subplot
  p_subplot <- p_subplot %>%
    layout(
      xaxis = list(title = if(i == 2) "Year" else ""),
      yaxis = list(title = "Indexed Value (2012 = 100)"),
      annotations = list(
        list(
          text = paste0("<b>", sector, "</b>"),
          x = 0.5,
          y = 1,
          xref = "paper",
          yref = "paper",
          xanchor = "center",
          yanchor = "top",
          showarrow = FALSE,
          font = list(size = 13, color = "black")
        )
      ),
      shapes = list(
        list(
          type = "line",
          x0 = 2012,
          x1 = 2023,
          xref = "x",
          y0 = 100,
          y1 = 100,
          yref = "y",
          line = list(color = "gray", width = 1, dash = "dot")
        )
      )
    )
  
  subplot_list_gu[[i]] <- p_subplot
}

# Combine subplots
p_govt_univ <- subplot(
  subplot_list_gu,
  nrows = 2,
  shareX = TRUE,
  titleY = TRUE,
  margin = 0.1
) %>%
  layout(
    title = list(
      text = "<b>Government and Higher Education R&D Value Added (Indexed to 2012 = 100)</b><br><sup>Critical components of the innovation ecosystem</sup>",
      font = list(size = 16)
    ),
    showlegend = TRUE,
    legend = list(
      orientation = "v",
      x = 1.02,
      y = 0.5,
      traceorder = "normal"
    ),
    margin = list(t = 120)
  )

p_govt_univ
```
*Note: "Higher Education" includes public and private nonprofit colleges and universities. "Government" includes federal, national defense, nondefense, and state and local governments.*

```{r rd_govt_univ_table}
#| tbl-cap: "Government and Higher Education R&D - All States (2023)"

rd_govt_univ_table <- rd_industry_detail %>%
  filter(sector_group %in% c("Government", "Higher Education"), year == 2023) %>%
  group_by(geography, sector_group) %>%
  summarise(value = sum(value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = sector_group, values_from = value) %>%
  mutate(
    across(c(Government, `Higher Education`), ~dollar(.x, suffix = "M", accuracy = 0.1)),
    sort_order = case_when(
      geography == "United States" ~ 1,
      geography == "Indiana" ~ 2,
      TRUE ~ 3
    )
  ) %>%
  arrange(sort_order) %>%
  select(State = geography, `Government R&D` = Government, `Higher Education R&D` = `Higher Education`) %>%
  select(-matches("sort_order"))

create_table(rd_govt_univ_table)
```

::: {.callout-note icon="false"}
## 🎓 Government and Higher Education R&D Insights
```{r govt_univ_insights}
#| output: asis

# Calculate key metrics
in_govt_2023 <- rd_industry_detail %>% filter(geography == "Indiana", sector_group == "Government", year == 2023) %>% pull(value)
in_univ_2023 <- rd_industry_detail %>% filter(geography == "Indiana", sector_group == "Higher Education", year == 2023) %>% pull(value)

in_govt_growth <- rd_industry_detail %>% 
  filter(geography == "Indiana", sector_group == "Government", year %in% c(2012, 2023)) %>%
  summarise(growth = ((last(value) / first(value)) - 1) * 100) %>% pull(growth)

in_univ_growth <- rd_industry_detail %>% 
  filter(geography == "Indiana", sector_group == "Higher Education", year %in% c(2012, 2023)) %>%
  summarise(growth = ((last(value) / first(value)) - 1) * 100) %>% pull(growth)

us_univ_growth <- rd_industry_detail %>% 
  filter(geography == "United States", sector_group == "Higher Education", year %in% c(2012, 2023)) %>%
  summarise(growth = ((last(value) / first(value)) - 1) * 100) %>% pull(growth)

# Get neighboring states data
neighbor_govt_2023 <- rd_industry_detail %>%
  filter(geography %in% c("Indiana", "Illinois", "Michigan", "Ohio", "Kentucky"), 
         sector_group == "Government", year == 2023) %>%
  arrange(desc(value))

neighbor_univ_2023 <- rd_industry_detail %>%
  filter(geography %in% c("Indiana", "Illinois", "Michigan", "Ohio", "Kentucky"), 
         sector_group == "Higher Education", year == 2023) %>%
  arrange(desc(value))

neighbor_govt_growth <- rd_industry_detail %>%
  filter(geography %in% c("Indiana", "Illinois", "Michigan", "Ohio", "Kentucky"), 
         sector_group == "Government", year %in% c(2012, 2023)) %>%
  group_by(geography) %>%
  summarise(growth = ((last(value) / first(value)) - 1) * 100, .groups = "drop") %>%
  arrange(desc(growth))

neighbor_univ_growth <- rd_industry_detail %>%
  filter(geography %in% c("Indiana", "Illinois", "Michigan", "Ohio", "Kentucky"), 
         sector_group == "Higher Education", year %in% c(2012, 2023)) %>%
  group_by(geography) %>%
  summarise(growth = ((last(value) / first(value)) - 1) * 100, .groups = "drop") %>%
  arrange(desc(growth))

# Get Indiana's ranks
in_govt_rank <- which(neighbor_govt_2023$geography == "Indiana")
in_univ_rank <- which(neighbor_univ_2023$geography == "Indiana")
in_govt_growth_rank <- which(neighbor_govt_growth$geography == "Indiana")
in_univ_growth_rank <- which(neighbor_univ_growth$geography == "Indiana")

# Helper for ordinal suffix
ordinal <- function(x) {
  suffix <- case_when(
    x == 1 ~ "st",
    x == 2 ~ "nd",
    x == 3 ~ "rd",
    TRUE ~ "th"
  )
  paste0(x, suffix)
}

cat(sprintf("- Indiana's government R&D totaled **$%sM** in 2023, growing **%.1f%%** since 2012\n", format(in_govt_2023, big.mark = ",", nsmall = 1), in_govt_growth))
cat(sprintf("- Higher education R&D reached **$%sM** in 2023, with **%.1f%%** growth (vs. **%.1f%%** nationally)\n", format(in_univ_2023, big.mark = ",", nsmall = 1), in_univ_growth, us_univ_growth))
cat(sprintf("- Among neighboring states, Indiana ranks **%s** in government R&D value added and **%s** in growth rate (%.1f%%)\n", 
            ordinal(in_govt_rank), ordinal(in_govt_growth_rank), in_govt_growth))
cat(sprintf("- For higher education R&D, Indiana ranks **%s** in total value and **%s** in growth (%.1f%%) among the five-state region\n", 
            ordinal(in_univ_rank), ordinal(in_univ_growth_rank), in_univ_growth))
cat("- University R&D is essential for **basic research**, **workforce development**, and **technology transfer** to industry\n")
cat("- States with strong university R&D systems often see spillover benefits in commercial innovation and startup formation\n")
```
:::

## Key R&D Insights
```{r rd_key_insights}
#| output: asis

# Calculate key metrics for callout
in_va_2023 <- rd_value_added %>% filter(geography == "Indiana", line_code == 1, year == 2023) %>% pull(value)
in_va_2012 <- rd_value_added %>% filter(geography == "Indiana", line_code == 1, year == 2012) %>% pull(value)
in_growth <- ((in_va_2023 / in_va_2012) - 1) * 100

us_va_2023 <- rd_value_added %>% filter(geography == "United States", line_code == 1, year == 2023) %>% pull(value)
in_va_share <- (in_va_2023 / us_va_2023) * 100

in_emp_2023 <- rd_employment %>% filter(geography == "Indiana", line_code == 1, year == 2023) %>% pull(value) * 1000
us_emp_2023 <- rd_employment %>% filter(geography == "United States", line_code == 1, year == 2023) %>% pull(value) * 1000
in_emp_share <- (in_emp_2023 / us_emp_2023) * 100

in_comp_2023 <- rd_compensation %>% filter(geography == "Indiana", line_code == 1, year == 2023) %>% pull(value)
us_comp_2023 <- rd_compensation %>% filter(geography == "United States", line_code == 1, year == 2023) %>% pull(value)
in_comp_share <- (in_comp_2023 / us_comp_2023) * 100

# Get Indiana's rank among neighboring states
in_rank <- rd_va_total %>% 
  filter(year == 2023, geography != "United States") %>% 
  arrange(desc(value)) %>% 
  mutate(rank = row_number()) %>% 
  filter(geography == "Indiana") %>% 
  pull(rank)

rank_suffix <- case_when(
  in_rank == 1 ~ "st",
  in_rank == 2 ~ "nd",
  in_rank == 3 ~ "rd",
  TRUE ~ "th"
)

# Sector breakdown insights
in_mfg_2023 <- rd_value_added %>% filter(geography == "Indiana", line_code == 4, year == 2023) %>% pull(value)
in_nonmfg_2023 <- rd_value_added %>% filter(geography == "Indiana", line_code == 9, year == 2023) %>% pull(value)
in_mfg_share <- (in_mfg_2023 / (in_mfg_2023 + in_nonmfg_2023)) * 100

in_mfg_growth <- rd_value_added %>%
  filter(geography == "Indiana", line_code == 4, year %in% c(2012, 2023)) %>%
  summarise(growth = ((last(value) / first(value)) - 1) * 100) %>%
  pull(growth)

in_nonmfg_growth <- rd_value_added %>%
  filter(geography == "Indiana", line_code == 9, year %in% c(2012, 2023)) %>%
  summarise(growth = ((last(value) / first(value)) - 1) * 100) %>%
  pull(growth)

# Get top manufacturing industry
top_mfg_industry <- rd_industry_detail %>%
  filter(geography == "Indiana", sector_group == "Manufacturing", year == 2023) %>%
  arrange(desc(value)) %>%
  slice(1)

# Gov't and Higher Ed
# Calculate metrics
in_govt_2023 <- rd_industry_detail %>% filter(geography == "Indiana", sector_group == "Government", year == 2023) %>% pull(value)
in_univ_2023 <- rd_industry_detail %>% filter(geography == "Indiana", sector_group == "Higher Education", year == 2023) %>% pull(value)

in_govt_growth <- rd_industry_detail %>% 
  filter(geography == "Indiana", sector_group == "Government", year %in% c(2012, 2023)) %>%
  summarise(growth = ((last(value) / first(value)) - 1) * 100) %>% pull(growth)

in_univ_growth <- rd_industry_detail %>% 
  filter(geography == "Indiana", sector_group == "Higher Education", year %in% c(2012, 2023)) %>%
  summarise(growth = ((last(value) / first(value)) - 1) * 100) %>% pull(growth)

# Get ranks
neighbor_univ_2023 <- rd_industry_detail %>%
  filter(geography %in% c("Indiana", "Illinois", "Michigan", "Ohio", "Kentucky"), 
         sector_group == "Higher Education", year == 2023) %>%
  arrange(desc(value))

in_univ_rank <- which(neighbor_univ_2023$geography == "Indiana")

univ_rank_suffix <- case_when(
  in_univ_rank == 1 ~ "st",
  in_univ_rank == 2 ~ "nd",
  in_univ_rank == 3 ~ "rd",
  TRUE ~ "th"
)

cat("**Overall R&D Performance:**\n\n")
cat(sprintf("- Indiana's R&D value added grew **%.1f%%** from 2012 to 2023, reaching **$%sM**\n", 
            in_growth, format(in_va_2023, big.mark = ",", nsmall = 1)))
cat(sprintf("- Indiana ranks **%d%s** among the five-state region in total R&D value added\n", in_rank, rank_suffix))
cat(sprintf("- Indiana accounts for **%.2f%%** of U.S. R&D value added, **%.2f%%** of R&D employment, and **%.2f%%** of R&D compensation\n", 
            in_va_share, in_emp_share, in_comp_share))
cat(sprintf("- **%s** workers were employed in R&D activities across Indiana in 2023\n\n", comma(in_emp_2023)))

cat("**Sector Composition & Dynamics:**\n\n")
cat(sprintf("- Manufacturing R&D dominates at **%.0f%%** of Indiana's total R&D value added, reflecting the state's industrial heritage\n", 
            in_mfg_share))
cat(sprintf("- Manufacturing R&D grew **%.1f%%** (2012-2023) while nonmanufacturing R&D grew **%.1f%%**—the gap is slightly **%s**\n",
            in_mfg_growth, in_nonmfg_growth, 
            if_else(in_mfg_growth > in_nonmfg_growth, "widening", "narrowing")))
cat(sprintf("- **%s** is Indiana's largest manufacturing R&D industry at **$%sM** in 2023\n",
            top_mfg_industry$description,
            format(top_mfg_industry$value, big.mark = ",", nsmall = 1)))
cat("- Compared to Ohio's balanced R&D portfolio, Indiana's persistent manufacturing dominance creates both **competitive advantages** in industrial innovation and **vulnerability** to sector-specific disruptions\n")

cat("\n**Government & Higher Education R&D:**\n\n")

cat(sprintf("- Government R&D totaled **$%sM** (%.1f%% growth) and higher education R&D reached **$%sM** (%.1f%% growth) from 2012-2023\n",
            format(in_govt_2023, big.mark = ",", nsmall = 1), in_govt_growth,
            format(in_univ_2023, big.mark = ",", nsmall = 1), in_univ_growth))
cat(sprintf("- Indiana ranks **%d%s** among neighboring states in higher education R&D value added\n", 
            in_univ_rank, univ_rank_suffix))
cat("- Public sector R&D — particularly university research — is critical for **basic research**, **talent development**, and **technology transfer** that supports private sector innovation\n")
cat("- States with robust university R&D ecosystems typically see stronger **startup formation** and **technology commercialization** outcomes\n")

```

## Data Notes

Key definitions from BEA R&D Satellite Account:

- **Value Added**: The contribution of R&D activities to GDP
- **Employment**: Full-time equivalent jobs in R&D activities
- **Compensation**: Total wages, salaries, and benefits for R&D workers
- **Five-State Region**: Indiana and its neighboring states (Michigan, Illinois, Ohio, and Kentucky)

See the full technical notes from BEA for detailed methodology.

# Labor Productivity Performance {#sec-labor-productivity}

Labor productivity, measured as output per hour worked, is a fundamental indicator of economic efficiency and competitiveness. This section examines Indiana's labor productivity trends relative to neighboring states using data from the [Bureau of Labor Statistics State Labor Productivity program](https://www.bls.gov/productivity/).

::: {.callout-note icon="false"}
## 📊 About Labor Productivity Indexes

The BLS labor productivity indexes measure **real output per hour worked** with 2017 as the base year (2017 = 100). These indexes:

- Capture efficiency gains from technological improvements, capital investment, and workforce skills
- Are adjusted for inflation to reflect real economic output
- Allow comparison across states and over time
- Reflect the combined productivity of all industries within a state's economy
:::

```{r productivity_data_import}
#| include: false

# Import labor productivity data
productivity_raw <- read_excel(file.path("CICP_20250709", "BLS Labor Productivity.xlsx"))

# Filter for five-state region and clean data
productivity_data <- productivity_raw %>%
  # Convert state FIPS to character for matching
  mutate(statefips = sprintf("%02d", statefips)) %>%
  # Filter for our five states
  filter(statefips %in% c("18", "17", "21", "26", "39")) %>%
  # Create clean state names
  mutate(
    state = case_when(
      statefips == "18" ~ "Indiana",
      statefips == "17" ~ "Illinois",
      statefips == "21" ~ "Kentucky",
      statefips == "26" ~ "Michigan",
      statefips == "39" ~ "Ohio",
      TRUE ~ area
    )
  ) %>%
  # Ensure year and data are numeric
  mutate(
    year = as.numeric(year),
    productivity_index = as.numeric(data)
  ) %>%
  # Keep only needed columns
  select(state, year, productivity_index) %>%
  # Remove any missing values
  filter(!is.na(productivity_index))

# Define color palette (matching other reports)
state_colors <- c(
  "Indiana" = "#1565C0",
  "Michigan" = "#D84315",
  "Illinois" = "#6A1B9A",
  "Ohio" = "#2E7D32",
  "Kentucky" = "#F57C00"
)
```

## Productivity Trends Over Time

The following chart shows how labor productivity has evolved across the five-state region from 2007-2024.
```{r productivity_trends}
#| fig-cap: "Labor Productivity Index: Indiana and Neighboring States (2017 = 100)"
#| fig-height: 9

# Create interactive line chart
p_productivity <- plot_ly()

# Add Indiana first (thicker line)
indiana_data <- productivity_data %>% filter(state == "Indiana")
p_productivity <- p_productivity %>%
  add_trace(
    data = indiana_data,
    x = ~year, y = ~productivity_index,
    name = "Indiana",
    type = 'scatter', mode = 'lines+markers',
    line = list(color = state_colors["Indiana"], width = 4),
    marker = list(color = state_colors["Indiana"], size = 8),
    hovertemplate = paste0(
      "<b>Indiana</b><br>",
      "Year: %{x}<br>",
      "Productivity Index: %{y:.1f}<br>",
      "<extra></extra>"
    )
  )

# Add other states
other_states <- c("Illinois", "Kentucky", "Michigan", "Ohio")
for(state in other_states) {
  state_subset <- productivity_data %>% filter(state == !!state)
  p_productivity <- p_productivity %>%
    add_trace(
      data = state_subset,
      x = ~year, y = ~productivity_index,
      name = state,
      type = 'scatter', mode = 'lines+markers',
      line = list(color = state_colors[state], width = 2),
      marker = list(color = state_colors[state], size = 6),
      hovertemplate = paste0(
        "<b>", state, "</b><br>",
        "Year: %{x}<br>",
        "Productivity Index: %{y:.1f}<br>",
        "<extra></extra>"
      )
    )
}

# Update layout
p_productivity <- p_productivity %>%
  layout(
    xaxis = list(
      title = "Year",
      gridcolor = "#E5E5E5",
      showgrid = TRUE
    ),
    yaxis = list(
      title = "Labor Productivity Index (2017 = 100)",
      gridcolor = "#E5E5E5",
      showgrid = TRUE
    ),
    hovermode = "x unified",
    plot_bgcolor = "white",
    paper_bgcolor = "white",
    font = list(family = "Arial, sans-serif", size = 12),
    legend = list(
      orientation = "h",
      yanchor = "bottom",
      y = -0.2,
      xanchor = "center",
      x = 0.5
    ),
    margin = list(b = 100)
  )

p_productivity
```

## Growth Rate Comparison

Comparing growth rates across different time periods reveals how productivity gains have been distributed over time.
```{r productivity_growth_table}
#| tbl-cap: "Labor Productivity Growth Rates by State and Time Period"

# Calculate growth rates for different periods
productivity_growth <- productivity_data %>%
  arrange(state, year) %>%
  group_by(state) %>%
  summarise(
    index_2007 = productivity_index[year == 2007],
    index_2024 = productivity_index[year == 2024],
    index_2017 = productivity_index[year == 2017],
    # Overall growth 2007-2024
    growth_total = ((index_2024 / index_2007) - 1) * 100,
    # Pre-base period 2007-2017
    growth_pre = ((index_2017 / index_2007) - 1) * 100,
    # Post-base period 2017-2024
    growth_post = ((index_2024 / index_2017) - 1) * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(growth_total))

# Create formatted table
productivity_growth_display <- productivity_growth %>%
  mutate(
    sort_order = case_when(
      state == "Indiana" ~ 0,
      TRUE ~ 1
    )
  ) %>%
  arrange(sort_order, desc(growth_total)) %>%
  select(
    State = state,
    `2007-2024` = growth_total,
    `2007-2017` = growth_pre,
    `2017-2024` = growth_post
  ) %>%
  mutate(
    across(where(is.numeric), ~sprintf("%.1f%%", .))
  )

create_table(productivity_growth_display, 
             title = "Cumulative labor productivity growth across time periods")
```

## Recent Performance (2020-2024)

The post-pandemic period offers insights into recent productivity dynamics.
```{r recent_productivity}
#| fig-cap: "Labor Productivity Growth: 2020-2024"
#| fig-height: 8

# Calculate 2020-2024 growth
recent_growth <- productivity_data %>%
  filter(year %in% c(2020, 2024)) %>%
  group_by(state) %>%
  summarise(
    growth = ((last(productivity_index) / first(productivity_index)) - 1) * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(growth)) %>%
  mutate(
    is_indiana = state == "Indiana",
    fill_color = ifelse(is_indiana, state_colors["Indiana"], "#95A5A6")
  )

# Create bar chart
p_recent <- plot_ly(
  data = recent_growth,
  x = ~growth,
  y = ~reorder(state, growth),
  type = 'bar',
  orientation = 'h',
  marker = list(color = ~fill_color),
  hovertemplate = paste0(
    "<b>%{y}</b><br>",
    "Growth: %{x:.1f}%<br>",
    "<extra></extra>"
  )
) %>%
  layout(
    xaxis = list(
      title = "Cumulative Growth (%)",
      gridcolor = "#E5E5E5",
      showgrid = TRUE
    ),
    yaxis = list(
      title = "",
      categoryorder = "total ascending"
    ),
    plot_bgcolor = "white",
    paper_bgcolor = "white",
    font = list(family = "Arial, sans-serif", size = 12),
    showlegend = FALSE,
    margin = list(l = 100)
  )

p_recent
```

::: {.callout-note icon="false"}
## 🔍 Key Productivity Insights
```{r productivity_insights}
#| output: asis

# Calculate key metrics
in_2024 <- productivity_data %>% filter(state == "Indiana", year == 2024) %>% pull(productivity_index)
in_2007 <- productivity_data %>% filter(state == "Indiana", year == 2007) %>% pull(productivity_index)
in_total_growth <- ((in_2024 / in_2007) - 1) * 100

in_2017 <- productivity_data %>% filter(state == "Indiana", year == 2017) %>% pull(productivity_index)
in_recent_growth <- ((in_2024 / in_2017) - 1) * 100

# Rankings
rank_2024 <- productivity_data %>%
  filter(year == 2024) %>%
  arrange(desc(productivity_index)) %>%
  mutate(rank = row_number()) %>%
  filter(state == "Indiana") %>%
  pull(rank)

rank_growth <- productivity_growth %>%
  arrange(desc(growth_total)) %>%
  mutate(rank = row_number()) %>%
  filter(state == "Indiana") %>%
  pull(rank)

# Get top and bottom performers
top_state <- productivity_growth %>% slice_max(growth_total, n = 1)
bottom_state <- productivity_growth %>% slice_min(growth_total, n = 1)

# Regional average (excluding Indiana)
regional_avg_growth <- productivity_growth %>%
  filter(state != "Indiana") %>%
  summarise(avg = mean(growth_total)) %>%
  pull(avg)

# Ordinal helper
ordinal <- function(x) {
  suffix <- case_when(
    x == 1 ~ "st",
    x == 2 ~ "nd", 
    x == 3 ~ "rd",
    TRUE ~ "th"
  )
  paste0(x, suffix)
}

cat(sprintf("- Indiana's labor productivity index reached **%.1f** in 2024, representing **%.1f%%** growth since 2007\n", 
            in_2024, in_total_growth))
cat(sprintf("- Compared to its neighbors, Indiana ranks **%s** in current productivity levels (2024)\n", 
            ordinal(rank_2024)))
cat(sprintf("- Indiana ranks **%s** in total productivity growth (2007-2024) among neighboring states\n", 
            ordinal(rank_growth)))
cat(sprintf("- **%s** leads the region with **%.1f%%** growth, while **%s** trails at **%.1f%%**\n",
            top_state$state, top_state$growth_total, bottom_state$state, bottom_state$growth_total))
cat(sprintf("- Indiana's growth of **%.1f%%** is **%s** the regional average of **%.1f%%** (excluding Indiana)\n",
            in_total_growth,
            ifelse(in_total_growth > regional_avg_growth, "above", "below"),
            regional_avg_growth))
cat(sprintf("- Since 2017, Indiana's productivity has grown **%.1f%%**, reflecting post-recession recovery and modernization efforts\n",
            in_recent_growth))
cat("- Labor productivity is influenced by **capital investment**, **technology adoption**, **workforce skills**, and **industry composition**\n")
cat("- States with higher productivity growth often see stronger **wage growth** and **economic resilience**\n")
```
:::

## Data Notes

**About the BLS State Labor Productivity Program:**

- Measures real output per hour worked for each state's private nonfarm business sector
- Base year is 2017 (2017 = 100), allowing for easy comparison across states and time periods
- Output is measured using real GDP data from the Bureau of Economic Analysis
- Hours worked data comes from the BLS Current Employment Statistics and Quarterly Census of Employment and Wages programs
- Productivity indexes reflect combined effects of capital investment, technology, workforce skills, and industrial composition

**Key Concepts:**

- **Labor Productivity**: Real economic output produced per hour of labor input
- **Index (2017=100)**: A value above 100 indicates productivity above the 2017 level; below 100 indicates productivity below the 2017 level
- **Growth Rates**: Percentage change in the index over a specified time period

For detailed methodology, see the [BLS State Labor Productivity Technical Notes](https://www.bls.gov/productivity/technical-notes/).